---
title: "R · Quarto setup"
subtitle: "Getting R, RStudio/Posit Cloud, and Quarto running for the simulation labs"
page-type: reference
toc: true
---

This page gets your software working so you can run the simulation code that appears in the labs. Read
it once near the start of the term, get a "hello, probability" report to render, and then come back to it
whenever something breaks.

One framing first, because it shapes everything below. In this course, **simulation supports probability
reasoning — it is not the center.** The mathematics is the point: a probability is a number you reason
about, and a simulation is a way to *check your reasoning* by letting the computer play out the random
process many times and counting. When you write that the shuttle is late with probability $0.19$, a short
simulation that comes back near $0.19$ is reassurance that you set the problem up correctly. It is never a
substitute for understanding *why* the number is $0.19$. So the software here stays deliberately small: a
few base-R functions, one file per task, and a habit of writing down what you expected before you run
anything.

A second framing about cost. **Everything you need is free and open.** This course does not use Cengage,
WebAssign, MyLab, or any paid statistics platform. R is free, the editors below are free (one of them runs
entirely in a browser with nothing to install), and Quarto is free. If a site ever asks you to pay to do
the coursework, that is not part of this class — check Blackboard and ask.

## What to install

You need three pieces. They stack: R does the computing, the editor is where you write and run code, and
Quarto turns a plain-text document with code in it into a clean report.

**1. R — the language that does the probability.**
R is the engine. Download it from CRAN, the Comprehensive R Archive Network, at
<https://cran.r-project.org>. Pick the build for your operating system (Windows, macOS, or Linux), install
it with the defaults, and you are done — you will rarely open R directly, because the editor below talks to
it for you. We use only **base R** in this course: the functions that ship with R itself, no add-on
packages to install. That keeps setup short and keeps the focus on the probability rather than on tooling.

**2. An editor — where you actually work.** Choose one of two. You do not need both.

- **RStudio Desktop** (recommended if you want everything on your own machine). It is a free, friendly
  front end for R: a place to type code, a console that runs it, a panel that shows your plots, and the
  one-button rendering you will use for reports. Download it from
  <https://posit.co/download/rstudio-desktop/> and install R first (step 1), then RStudio — RStudio finds
  your R installation automatically.
- **Posit Cloud** (recommended if you would rather not install anything). It is RStudio running in your web
  browser, hosted by Posit. Create a free account at <https://posit.cloud>, start a new project, and you
  have R, the editor, and Quarto already set up and connected — nothing to download. This is the easiest
  path if you are on a managed or borrowed computer, and the free tier is enough for this course's small
  simulations.

Either choice gives you the same experience for our purposes. Pick the one that fits your machine and move
on.

**3. Quarto — the document system that makes the report.** Quarto takes a plain-text file (a `.qmd` file)
that mixes your writing and your code, runs the code, and produces a tidy HTML or PDF report with the
results and figures dropped in. Get it from <https://quarto.org/docs/get-started/>. **If you installed a
current RStudio Desktop, Quarto came bundled** and you can skip the separate download; install it on its own
only if `quarto check` (below) says it is missing. On Posit Cloud, Quarto is already there.

## Make R discoverable by Quarto

"Discoverable" just means Quarto can find your R installation so it can run the code in your document. In
the common setups this is automatic, but here is how to confirm it and what to do if it is not.

- **Working in RStudio Desktop or Posit Cloud (the usual case).** Open your `.qmd` file and click the
  **Render** button at the top of the editor. RStudio hands the file to Quarto, Quarto runs the R code
  through the R that RStudio already knows about, and a preview appears. If a report renders, R is
  discoverable — you are set, and you can stop here.
- **Confirming from a terminal.** Open a terminal (in RStudio: the *Terminal* tab next to the Console) and
  run `quarto check`. It reports the Quarto version, whether it found R, and whether the small support
  package it needs for R is in place. Green checks mean you are ready. If it reports a missing piece, the
  message names it, and on RStudio the fix is usually to let it install the support package when prompted.
- **Posit Cloud is pre-connected.** A new Posit Cloud project already has R, the editor, and Quarto wired
  together, so `quarto check` there is mostly a sanity confirmation rather than a fix-it step.

If you are stuck on this step, do not spend an evening on it alone — bring it to office hours or the class
help channel. Getting one report to render is the only setup milestone that matters.

## Your first reproducible report

Here is a minimal Quarto document. Create a new file called `hello-probability.qmd`, paste this in, and
render it. It estimates the chance the shuttle is **late**, which in our running "commuter's morning"
example is $P(\text{late}) = 0.19$ (these are synthetic numbers, seed set), and checks that a simulation
lands nearby.

The block below is shown as text so you can read it as a whole; copy it into your own `.qmd` and render
there.

````markdown
---
title: "Hello, probability"
format: html
---

We model each weekday as a single trial that is "late" with probability 0.19,
then simulate a term's worth of mornings and compare the simulated rate to 0.19.

```{r}
#| label: simulate-late-shuttle
#| eval: false
set.seed(35003)
days   <- 75                         # mornings in a term
late   <- rbinom(n = days, size = 1, prob = 0.19)   # 1 = late, 0 = on time
c(simulated_late_rate = mean(late), expected = 0.19)
```
````

A few things to understand, because you will see this pattern in every lab.

- **`#| label: simulate-late-shuttle`** gives the code chunk a name. Labels make error messages point at
  the right place and let you refer to a result later; get in the habit of naming every chunk something
  descriptive.
- **`#| eval: false`** tells Quarto **not** to run this chunk when *this site* is built — that is why the
  code on the course pages is shown but never executed here. **In your own copy you set it to
  `#| eval: true`** (or simply delete the line, since chunks run by default) so the code actually runs and
  you see the result. Reading code is not the same as running it; you learn the probability by running it.
- **`set.seed(35003)`** fixes R's random-number generator to a known starting point. `rbinom` produces
  "random" draws from a deterministic stream, and seeding it means *you* and *anyone else who runs the same
  file* get the *same* draws — so the report is reproducible. `rbinom(n = days, size = 1, prob = 0.19)`
  asks for `days` independent yes/no trials, each late with probability $0.19$; `mean()` of those zeros and
  ones is the simulated late rate, which should sit near $0.19$.

When you render your copy with `#| eval: true`, the simulated rate will be close to $0.19$ but not exactly
$0.19$ — that gap *is* the lesson. A simulation estimates a probability; it does not replace the exact
calculation. The exact value comes from the model, and we keep the simulation as a check.

## Reproducibility habits we keep all term

These three habits make your work trustworthy — to your future self, to a classmate, and to anyone
reading your report. They cost almost nothing and they pay off the first time something looks surprising.

- **Always `set.seed(35003)` before any simulation.** We use the same seed across the whole course so that
  the numbers in your session match the numbers discussed in class and in the notes. If you change the
  seed, your draws change, and your "0.19-ish" result will be a *different* 0.19-ish result — fine in
  principle, confusing when you are comparing notes. Seed first, simulate second, every time.
- **End each report with `sessionInfo()`.** This prints your R version and platform, so if a result cannot
  be reproduced later, the record of *what you ran it on* is right there in the report. Include it as a
  final chunk (shown here, as everywhere on this site, with `#| eval: false`; in your copy let it run):

```{r}
#| label: session-info
#| eval: false
sessionInfo()
```

- **One file per task.** Keep each lab or exploration in its own `.qmd` with a clear name
  (`hello-probability.qmd`, `lab-02-monte-carlo.qmd`, and so on). One file, one question, one set of
  results. It is far easier to debug, to re-run, and to hand in than a single sprawling file that does
  five things. When a task is done, its file stands on its own and renders from top to bottom without you
  having to remember a hidden step.

Together these habits mean any report on your machine can be re-run by you next month, or by a classmate
on a different computer, and produce the same numbers. That reproducibility is the whole reason we bother
with code in a probability course.

## AI Use Note (when you use an assistant)

You may use an AI assistant (a chatbot or a coding helper) to get unstuck — to explain an error message,
suggest why a chunk will not render, or remind you what an R function does. The expectation is simple:
**you remain responsible for everything in your file, and you verify what the tool gives you.** An
assistant can produce code that runs cleanly and is still wrong about the probability — a simulation that
estimates the wrong quantity renders perfectly. So treat assistant output the way you treat your own first
draft: check it against the model and against a simulation before you trust it.

When you lean on an assistant, jot down a one-line record of how. Capturing it keeps you honest and makes
it easy to explain your process if asked.

| Field | What it captures |
|---|---|
| **Tool** | Which assistant you used (name and, if relevant, the version). |
| **Purpose** | What you asked it for — e.g. "explain a Quarto render error," "remind me of `rbinom`'s arguments." |
| **Verification** | **How you confirmed the result is correct** — the load-bearing part: which exact calculation or simulation you checked it against, and whether it agreed. Output you did not verify is output you cannot rely on. |

The Verification row is the one that matters. If you cannot say how you checked an assistant's answer, you
do not yet know whether it is right, and you should not put it in your report.

## Public vs. graded

These notes, the examples, and the practice here are **public and ungraded** — study material only. No
graded prompts, answer keys, rubrics, point values, or due dates appear on this site. Graded checkpoints,
quizzes, homework, labs, the midterm, the project, and the final live in **Blackboard (the LMS)**, which is
authoritative for due dates, submissions, and grades. If this page and Blackboard ever disagree, follow
Blackboard.

## See also

- [Labs overview](../labs/index.qmd) — the simulation labs where you put this setup to work.
- [Notation glossary](notation-glossary.qmd) — the symbols used across the course, including
  $P(\text{late})$ and the distribution names you will pass to R's random-number functions.
- [Distribution reference](distribution-reference.qmd) — the standard models and the exact
  parameterizations this course uses, which match the arguments you give functions like `rbinom`.
