---
title: "Lab 2 — Monte Carlo basics"
subtitle: "Estimating a probability by simulating it"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 2 — Sample spaces, events & rules](../notes/week-02-sample-spaces-events-rules.qmd). There you
> built the shuttle×rain sample space and computed $P(\text{on time}) = 0.81$ by hand using the addition
> and complement rules. Here you will reach the *same* number a completely different way: by simulating
> Maya's morning thousands of times and counting how often the shuttle is on time. When the hand
> calculation and the simulation agree, you trust both more — and you get a first, concrete taste of the
> **law of large numbers**.

## The idea

A probability is a long-run proportion. When we say $P(\text{on time}) = 0.81$, one honest reading is:
"if Maya took this commute on a very large number of independent mornings, the fraction of mornings the
shuttle arrived on time would settle near $0.81$." That sentence is not just an interpretation — it is a
recipe. If we can *act out* one morning at random on the computer, we can act out ten thousand mornings,
tally the on-time ones, and divide. The resulting fraction is a **Monte Carlo estimate** of the
probability.

Monte Carlo is the name for this whole family of methods: use a stream of random numbers to imitate a
random process, then read an answer off the imitation. It is powerful precisely when the hand
calculation is hard or impossible. The shuttle problem is *not* hard — we already know the exact answer
is $0.81$ — and that is exactly why it makes a good first lab. We can check the simulation against the
truth and learn to trust the technique before we point it at problems we cannot solve by hand.

The world we are simulating is the same locked synthetic world from the week note. Each morning:

- It rains with probability $P(\text{rain}) = 0.30$.
- If it rains, the shuttle is on time with probability $P(\text{on time}\mid\text{rain}) = 0.60$.
- If it does not rain, the shuttle is on time with probability $P(\text{on time}\mid\text{no rain}) = 0.90$.

Averaging over the weather gives the marginal probability you derived by hand:

$$
P(\text{on time}) = 0.60 \times 0.30 + 0.90 \times 0.70 = 0.81,
\qquad
P(\text{late}) = 1 - 0.81 = 0.19.
$$

All numbers here are **synthetic; seed set.** Nothing depends on real weather or a real shuttle.

## Goal

By the end of this lab you should be able to:

- Translate a two-stage random process (weather, then shuttle) into a short base-R simulation.
- Estimate $P(\text{on time})$ and $P(\text{late})$ as the mean of a $0/1$ indicator over many
  simulated mornings.
- Compare your simulated estimate to the exact value $0.81$ and judge whether the gap is "small."
- Watch the running estimate **settle** toward $0.81$ as the number of replications grows — the law of
  large numbers in action.
- Change the inputs (the rain and on-time probabilities) and re-estimate, confirming the simulation
  tracks whatever world you specify.

This is self-check, ungraded practice. There is nothing to submit here; work the steps, predict each
result before you imagine running the chunk, and notice where your prediction and the logic agree.

## Setup

You need R and the ability to open a Quarto (`.qmd`) or plain R script — RStudio or Posit Cloud both
work. No add-on packages are required: everything below is **base R**. The functions we lean on are
small and worth knowing by name:

- `set.seed()` — fixes the random-number stream so the "random" results are reproducible. Every chunk
  that draws randomness in this lab starts with `set.seed(35003)` (the course seed) so your numbers match
  the discussion.
- `rbinom(n, size, prob)` — draws `n` values, each the count of successes in `size` independent trials
  with success probability `prob`. With `size = 1` it is a coin flip returning `0` or `1`, which is
  exactly what we want for a yes/no event like "did it rain?"
- `ifelse(test, yes, no)` — a vectorized choice: wherever `test` is `TRUE` it returns `yes`, elsewhere
  `no`. We use it to give each morning the right on-time probability depending on whether it rained.
- `mean()` — of a vector of `0`s and `1`s, this is just the proportion of `1`s, i.e. our estimated
  probability.

::: {.callout-note}
Here the code is **shown, not run**. Each chunk is marked `#| eval: false`, so the page
renders the code as teaching without executing it. To actually see the numbers, copy a chunk into your
own R session. The chunks are written so that, when executed, the seed makes them reproducible.
:::

## Steps

We build the simulation in pieces: first generate the weather, then the shuttle outcome conditional on
the weather, then collapse everything to a single estimate, and finally watch that estimate stabilize.

### Step 1 — Simulate the weather for many mornings

Pick a number of mornings to simulate — call it `n_morn`. Start large enough that the estimate is stable
(ten thousand is plenty here) but small enough to run instantly. Each morning, rain is a single yes/no
trial with probability $0.30$, so `rbinom(n_morn, size = 1, prob = 0.30)` gives us a vector of `n_morn`
values, each `1` for rain and `0` for no rain. The mean of that vector should land near $0.30$, a tiny
first check that the draw did what we asked.

```{r}
#| label: step1-simulate-rain
#| eval: false

set.seed(35003)            # reproducible: course seed

n_morn <- 10000            # number of simulated mornings
p_rain <- 0.30             # P(rain), locked synthetic value

# Draw rain for each morning: 1 = rain, 0 = no rain
rain <- rbinom(n_morn, size = 1, prob = p_rain)

# Sanity check: fraction of rainy mornings should sit near 0.30
mean(rain)
```

The object `rain` is now a length-`10000` vector of `0`s and `1`s. We have *not* touched the shuttle
yet — we have only rolled the weather. Keeping the two stages separate mirrors exactly how you built the
sample space by hand: branch on the weather first, then on the shuttle.

### Step 2 — Simulate the shuttle, conditional on the weather

The shuttle's on-time probability is **not** the same every morning; it depends on whether it rained.
This conditional structure is the whole point of the week, so the code makes it explicit. For each
morning we want the success probability to be $0.60$ on rainy mornings and $0.90$ on dry ones. The
`ifelse()` function builds exactly that vector of per-morning probabilities, and then a single call to
`rbinom()` with a *vector* of probabilities draws one shuttle outcome per morning, each with its own
correct probability.

```{r}
#| label: step2-simulate-ontime
#| eval: false

# Per-morning on-time probability depends on rain:
#   rain == 1  -> 0.60
#   rain == 0  -> 0.90
p_ontime <- ifelse(rain == 1, 0.60, 0.90)

# One shuttle outcome per morning: 1 = on time, 0 = late
on_time <- rbinom(n_morn, size = 1, prob = p_ontime)

# Peek at the first few mornings (rain, its on-time prob, the outcome)
head(data.frame(rain, p_ontime, on_time))
```

Read that `head()` output as a tiny logbook of Maya's first six simulated mornings: the weather, the
probability the shuttle "rolled against," and whether it actually made it on time. Notice that even on a
rainy morning the shuttle can be on time, and even on a dry morning it can be late — the randomness lives
in the `rbinom` draw, not in a hard rule.

### Step 3 — Estimate P(on time) and P(late)

Now collapse the whole simulation to a number. The vector `on_time` holds a `1` for every on-time
morning and a `0` for every late one, so its **mean is the proportion of on-time mornings** — our Monte
Carlo estimate of $P(\text{on time})$. The estimate of $P(\text{late})$ is one minus that, or
equivalently the mean of the late indicator. Treating an event's $0/1$ indicator and then averaging is
the core move of the whole lab; everything else is bookkeeping.

```{r}
#| label: step3-estimate-prob
#| eval: false

# Estimated P(on time): proportion of mornings the shuttle was on time
p_ontime_hat <- mean(on_time)

# Estimated P(late): the complement (mean of the late indicator)
p_late_hat <- mean(on_time == 0)

p_ontime_hat
p_late_hat
p_ontime_hat + p_late_hat   # should be exactly 1: the two events partition every morning
```

With the course seed and `n_morn <- 10000`, you should get an on-time estimate close to (but not exactly
equal to) $0.81$ — typically within about a percentage point. The last line is a structural check, not a
random one: on-time and late are complementary events, so their estimates must sum to exactly $1$ no
matter what the simulation produced.

### Step 4 — Watch the estimate settle (a first look at the LLN)

A single estimate from ten thousand mornings is reassuring, but the deeper lesson is *how* the estimate
behaves as the number of mornings grows. The **law of large numbers** says the running proportion of
on-time mornings converges to the true probability $0.81$ as we accumulate more and more mornings. We can
see this directly: form the **running mean** of `on_time` — the estimate using just the first morning,
then the first two, then the first three, and so on — using `cumsum()` divided by the running count.
Early on the running estimate lurches around; as the morning count climbs it tightens toward $0.81$.

```{r}
#| label: step4-running-estimate-lln
#| eval: false
#| fig-cap: "Running Monte Carlo estimate of P(shuttle on time) settling toward the exact value 0.81 as more mornings are simulated."
#| fig-alt: "A line that swings widely for the first few mornings then flattens out near a dashed reference line at 0.81 as the number of simulated mornings grows."

set.seed(35003)            # re-seed so this block stands alone

n_morn <- 10000
rain    <- rbinom(n_morn, size = 1, prob = 0.30)
on_time <- rbinom(n_morn, size = 1, prob = ifelse(rain == 1, 0.60, 0.90))

# Running estimate of P(on time) after 1, 2, 3, ... mornings
running_est <- cumsum(on_time) / seq_len(n_morn)

# Compare the estimate at a few checkpoints to the exact value 0.81
checkpoints <- c(10, 100, 1000, 10000)
data.frame(mornings = checkpoints,
           estimate = round(running_est[checkpoints], 3),
           exact    = 0.81)

# A simple base-R picture of the settling behavior
plot(running_est, type = "l",
     xlab = "number of simulated mornings",
     ylab = "running estimate of P(on time)",
     main = "Estimate settling toward 0.81")
abline(h = 0.81, lty = 2)   # the exact value, for reference
```

The data frame makes the settling concrete: at 10 mornings the estimate can be far off, but by 1,000 and
especially 10,000 mornings it hugs $0.81$. The plot tells the same story as a picture — a jagged curve
early, flattening toward the dashed line at $0.81$. That convergence *is* the law of large numbers, and
it is the reason a long simulation can stand in for an exact probability when no exact answer is at hand.

### Step 5 — Change the world and re-estimate (hands-on)

The simulation tracks whatever world you specify, not just the locked one. Change the inputs and the
estimate moves to match. Suppose you wanted to explore a rainier town where the shuttle is also a bit
worse in the rain: raise $P(\text{rain})$ to $0.50$ and drop $P(\text{on time}\mid\text{rain})$ to
$0.50$, keeping the dry-day on-time probability at $0.90$. Before running it, predict the new marginal by
hand — $0.50 \times 0.50 + 0.90 \times 0.50 = 0.70$ — then let the simulation confirm it.

```{r}
#| label: step5-change-the-world
#| eval: false

set.seed(35003)

n_morn <- 10000

# New, made-up world (synthetic): rainier, worse shuttle in the rain
p_rain_new        <- 0.50
p_ontime_rain_new <- 0.50
p_ontime_dry_new  <- 0.90

rain_new    <- rbinom(n_morn, size = 1, prob = p_rain_new)
on_time_new <- rbinom(n_morn, size = 1,
                      prob = ifelse(rain_new == 1, p_ontime_rain_new, p_ontime_dry_new))

p_ontime_hat_new <- mean(on_time_new)

# Hand-computed marginal for this new world, to compare against
exact_new <- p_ontime_rain_new * p_rain_new + p_ontime_dry_new * (1 - p_rain_new)

c(estimate = round(p_ontime_hat_new, 3), exact = exact_new)
```

The simulated estimate should land near $0.70$, matching your hand calculation for the new world. The
takeaway: the *machinery* is identical — draw the weather, draw the shuttle conditional on the weather,
average the indicator. Only the inputs changed. That is what makes Monte Carlo reusable.

## Verify

Verification means asking, deliberately, whether the simulation agrees with something you already trust.

1. **Compare to the exact value.** For the locked world, the hand answer from Week 2 is
   $P(\text{on time}) = 0.81$. Your Step 3 estimate from ten thousand mornings should differ from $0.81$
   by only a small amount — on the order of a percentage point or less. If it were off by, say, $0.10$,
   something is wrong: re-check that the conditional probabilities are wired to the *right* weather
   (`0.60` for rain, `0.90` for no rain, not swapped).
2. **Check the complement.** $\hat P(\text{on time}) + \hat P(\text{late})$ must equal exactly $1$. This
   is structural, so it is a check on your code rather than on the randomness.
3. **Check the weather draw.** `mean(rain)` should sit near $0.30$. If it does not, the rain probability
   is misspecified and every later number inherits the error.
4. **Read the settling.** In Step 4, the estimate at $10{,}000$ mornings should be visibly closer to
   $0.81$ than the estimate at $10$ mornings. If the running curve is *not* tightening toward the dashed
   line, suspect that the seed or the draw is being reset inside the loop.
5. **Re-derive the changed world.** In Step 5, the simulation estimate and the hand-computed `exact_new`
   should agree to within simulation noise. Agreement across a *different* set of inputs is stronger
   evidence the method is right than a single matching number.

A simulated estimate is never expected to hit the exact value *precisely* — that small leftover gap is
ordinary sampling variability, and it shrinks as you simulate more mornings. The goal is agreement
"within noise," not to the last decimal.

## AI use note

An AI assistant can help you learn the base-R idioms in this lab, but it cannot certify that your
simulation is correct — only comparison against the exact value $0.81$ and your own hand calculations can
do that. Use AI to get unstuck, then verify independently. Disclose any AI help per the course policy.

| Tool | Purpose | Verification |
|------|---------|--------------|
| Chat assistant (e.g. an LLM) | Explain what `rbinom(n, size = 1, prob)` returns and why `size = 1` gives a yes/no draw | Re-read R's `?rbinom`; confirm `mean(rain)` is near $0.30$ |
| Chat assistant | Suggest the `ifelse()` pattern for per-morning conditional probabilities | Inspect the `head()` logbook: rainy rows must show prob `0.60`, dry rows `0.90` |
| Chat assistant | Help read a base-R error message or fix a typo in a chunk | Run the corrected chunk; check the estimate lands near $0.81$ |
| Code completion tool | Draft the running-mean / `cumsum()` line in Step 4 | Confirm the curve settles toward the dashed line at $0.81$ |

The non-negotiable rule: whatever an AI tool writes, you still verify the answer against the exact
probability and the complement check. The simulation is a witness, not an authority — the hand-derived
$0.81$ is the standard it must meet.

## See also

- [Week 2 — Sample spaces, events & rules](../notes/week-02-sample-spaces-events-rules.qmd) — the hand
  derivation of $P(\text{on time}) = 0.81$ this lab reproduces by simulation.
- [Week 1 — Uncertainty, probability & models](../notes/week-01-uncertainty-probability-models.qmd) —
  what the number $0.81$ *means* as a long-run proportion.
- [Week 13 — Sums, simulation & limit behavior](../notes/week-13-sums-simulation-limit-behavior.qmd) —
  the law of large numbers and the central limit theorem, where the settling you saw in Step 4 becomes
  the headline.
- [Lab 5 — Bayes by simulation](lab-05-bayes-by-simulation.qmd) — the next simulation lab, reversing the
  conditioning to estimate $P(\text{rain}\mid\text{late})$.
- [Notation glossary](../resources/notation-glossary.qmd) — symbols and parameterizations used above.

*The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.*
