---
title: "Lab 9 — Simulating discrete models"
subtitle: "Binomial and Poisson by simulation, checked against the formulas"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 9 — Common discrete models](../notes/week-09-common-discrete-models.qmd). There you derived two
> probabilities by hand: that a pure guesser scores eight or better on the ten-question quiz about
> $5.5\%$ of the time, and that exactly four shuttles arrive in an hour about $20\%$ of the time. Here you
> *generate* those situations many times in R, count what actually happens, and check that the simulated
> answers land on the formulas. When they agree, you have two independent confirmations of the same
> result — and a working test of whether you understood the model.

## The idea

The notes give you two ways to answer a probability question, and this lab is about watching them meet. The
first way is the **formula**: write down the pmf, plug in, and compute a single exact number — for the
quiz, $P(X \ge 8) = 56/1024 \approx 0.0547$; for the shuttle, $P(N = 4) = e^{-4}4^4/4! \approx 0.195$. The
second way is **simulation**: build a machine that produces the random outcome — a guessed quiz, an hour of
arrivals — run it thousands of times, and count the fraction of runs that fall in the event you care
about. That fraction is an *estimate* of the probability.

These two answers are not the same kind of thing. The formula is exact and fixed; the simulated fraction
is random and will wobble a little each time you change the seed or the number of runs. But the wobble
shrinks as you do more runs (that is the law of large numbers, which Week 13 makes precise), and for the
sample sizes here the estimate should sit close to the formula. The teaching point is exactly that
**agreement**: if your simulation and your formula give nearly the same number, each one corroborates the
other. If they disagree badly, you have a bug — in the code, or in your understanding of the model — and
the gap is telling you where to look. Simulation here is not a replacement for the probability; it is a
check on it.

A second, quieter idea runs underneath: a named model is a *recipe for generating data*. Saying
"$X \sim \text{Binomial}(10, 0.5)$" is the same as saying "flip ten fair coins and count heads," and R has
a function (`rbinom`) that does exactly that flipping for you. Once you see a distribution as a generator,
simulation stops being mysterious — you are just running the generator and tallying.

## Goal

By the end of this lab you should be able to:

- Generate many draws from a $\text{Binomial}(10, 0.5)$ model with `rbinom`, tabulate the empirical pmf,
  and compare it bar-for-bar with the exact pmf from `dbinom`.
- Estimate $P(X \ge 8)$ for the guessing quiz by simulation and confirm it lands near the formula value
  $\approx 0.0547$.
- Generate many draws from a $\text{Poisson}(4)$ model with `rpois` and estimate $P(N = 4)$, confirming it
  lands near $\approx 0.195$.
- Explain *why* the simulated fraction is close to but not exactly equal to the formula, and what would
  make it closer.

You are not turning anything in here. This is study and self-check: the value is in reading each line of
code, predicting what it will print, and then confirming.

## Setup

You need R (with RStudio or Posit Cloud) and nothing beyond base R — no packages to install. If you have
not set up your environment yet, the [R · Quarto setup](../resources/r-quarto-setup.qmd) page walks through
it. Open a fresh script or a `.qmd` document and work through the steps in order.

Three things to fix before you start, so your numbers match the discussion:

- **The seed.** Every simulation chunk in this lab begins with `set.seed(35003)`. The seed makes the
  "random" draws reproducible: you and a classmate running the same code get the same numbers. Change the
  seed and the estimates shift slightly — that wobble is the point of the Verify section, not a mistake.
- **The number of runs.** We use `n_sim <- 100000` (one hundred thousand) draws. More runs means a tighter
  estimate; one hundred thousand is plenty to pin both probabilities to three decimals' worth of
  confidence without making your laptop wait.
- **The models, from the notes.** The quiz is $X \sim \text{Binomial}(10,\ 0.5)$ — ten true/false
  questions, pure guessing, count the number correct. The shuttle is $N \sim \text{Poisson}(4)$ — arrivals
  in an hour at a rate of four per hour. Both worlds and all numbers are **synthetic and fixed for the
  course**.

A reminder on the code blocks below: they are **shown for study**, written in base R with `#| eval: false`,
so this site does not run them. You run them yourself in your own session and read the printed output
there.

## Steps

### Step 1 — Simulate the guessing quiz and build the empirical pmf

A single guessed quiz is ten independent fair coin flips with "correct" counted as a success. One call to
`rbinom(n, size, prob)` draws `n` such quizzes at once: `size = 10` questions, `prob = 0.5` chance correct.
The result is a vector of `n_sim` whole numbers, each between $0$ and $10$ — the score on one simulated
quiz. We tabulate those scores with `table()`, divide by the number of quizzes to turn counts into
*proportions*, and that proportion vector is the **empirical pmf**: the simulated stand-in for $p(x)$.

```{r}
#| label: sim-binom-empirical-pmf
#| eval: false

set.seed(35003)

n_sim <- 100000                 # number of simulated quizzes
n_q   <- 10                     # questions per quiz
p     <- 0.5                    # chance of a correct guess

# Draw n_sim quizzes; each value is the number correct on that quiz (0..10).
scores <- rbinom(n_sim, size = n_q, prob = p)

# Empirical pmf: the fraction of quizzes landing on each score 0..10.
emp_counts <- table(factor(scores, levels = 0:n_q))
emp_pmf    <- emp_counts / n_sim

round(emp_pmf, 4)               # the simulated p(x), one entry per score
```

Wrapping `scores` in `factor(..., levels = 0:n_q)` forces the table to show all eleven possible scores,
even any that happened zero times (a score of $0$ or $10$ is rare and might be missing from a smaller run).
Read the printed vector as a probability distribution: most of the mass should pile up around $5$, the
mean, and thin out toward the ends.

### Step 2 — Overlay the exact pmf from the formula

Now generate the *exact* pmf — the formula's answer — and set it beside the simulated one. The function
`dbinom(x, size, prob)` returns $p(x) = \binom{10}{x}(0.5)^{x}(0.5)^{10-x}$ for each `x` you ask about, so
`dbinom(0:10, 10, 0.5)` gives the whole pmf in one line. Stacking the empirical and exact vectors into a
small table lets you compare them score by score and eyeball the gap.

```{r}
#| label: sim-binom-overlay-exact
#| eval: false

set.seed(35003)

n_sim <- 100000
n_q   <- 10
p     <- 0.5

scores  <- rbinom(n_sim, size = n_q, prob = p)
emp_pmf <- as.numeric(table(factor(scores, levels = 0:n_q))) / n_sim

# Exact pmf straight from the binomial formula.
exact_pmf <- dbinom(0:n_q, size = n_q, prob = p)

# Put them side by side and look at the difference column.
compare <- data.frame(
  score      = 0:n_q,
  empirical  = round(emp_pmf,   4),
  exact      = round(exact_pmf, 4),
  difference = round(emp_pmf - exact_pmf, 4)
)
print(compare)
```

The `difference` column is where the lesson lives. Each entry should be small — a few thousandths at most —
and the differences should scatter around zero rather than all leaning one way. That scatter is sampling
noise: with one hundred thousand quizzes the empirical pmf tracks the formula closely but not perfectly. If
you also want a picture, `barplot(rbind(emp_pmf, exact_pmf), beside = TRUE)` draws the two pmfs as paired
bars so the overlay is visible at a glance.

### Step 3 — Estimate the tail probability P(X ≥ 8)

The headline number from the notes is $P(X \ge 8)$, the chance a pure guesser scores eight or more. You do
*not* need the table for this — you can ask the original `scores` vector directly. The expression
`scores >= 8` produces a vector of `TRUE`/`FALSE`, one per quiz; in R, `TRUE` counts as $1$ and `FALSE` as
$0$, so `mean(scores >= 8)` is the *fraction* of quizzes scoring eight or better. That fraction is the
Monte Carlo estimate of the probability.

```{r}
#| label: sim-binom-tail-prob
#| eval: false

set.seed(35003)

n_sim <- 100000
scores <- rbinom(n_sim, size = 10, prob = 0.5)

# Fraction of simulated quizzes scoring 8, 9, or 10 correct.
p_hat_tail <- mean(scores >= 8)

# Exact value from the formula, for comparison.
p_exact_tail <- sum(dbinom(8:10, size = 10, prob = 0.5))   # = pbinom(7, 10, 0.5, lower.tail = FALSE)

cat("Simulated P(X >= 8):", round(p_hat_tail,   4), "\n")
cat("Exact     P(X >= 8):", round(p_exact_tail, 4), "\n")
```

The exact value sums the top three pmf bars — $\binom{10}{8} + \binom{10}{9} + \binom{10}{10} = 45+10+1 =
56$ orderings over $1024$ — giving $56/1024 \approx 0.0547$. Your simulated `p_hat_tail` should print
something very near it, off only in the third or fourth decimal.

### Step 4 — Simulate shuttle arrivals and estimate P(N = 4)

Switch worlds to the shuttle. Arrivals in an hour follow $N \sim \text{Poisson}(4)$, and `rpois(n, lambda)`
generates `n` such hourly counts with `lambda = 4`. To estimate $P(N = 4)$ — exactly four arrivals — count
the fraction of simulated hours equal to four with `mean(arrivals == 4)`, the same `TRUE`-as-$1$ trick as
before. Then compare to the formula via `dpois(4, 4)`.

```{r}
#| label: sim-pois-equals-four
#| eval: false

set.seed(35003)

n_sim  <- 100000
lambda <- 4                     # arrivals per hour

# Draw n_sim hours; each value is the number of shuttles that arrived that hour.
arrivals <- rpois(n_sim, lambda = lambda)

# Estimated and exact P(N = 4).
p_hat_four   <- mean(arrivals == 4)
p_exact_four <- dpois(4, lambda = lambda)         # e^-4 * 4^4 / 4!

cat("Simulated P(N = 4):", round(p_hat_four,   4), "\n")
cat("Exact     P(N = 4):", round(p_exact_four, 4), "\n")

# Sanity check on the model: a Poisson's mean and variance both equal lambda.
cat("sample mean:", round(mean(arrivals), 3),
    " sample var:", round(var(arrivals), 3), "\n")
```

The exact value is $e^{-4}4^4/4! \approx 0.195$, and `p_hat_four` should land beside it. The last two lines
are a bonus check on the *whole* model rather than one probability: a Poisson has mean equal to variance,
both $\lambda = 4$, so your `sample mean` and `sample var` should both come out near $4$. If they were far
apart, that would warn you the data are not really Poisson.

## Verify

This is the section the whole lab points at: line up each simulated estimate against the exact value from
the formula and confirm they agree to within sampling noise.

| Quantity | Formula (exact) | Simulated estimate (seed 35003, 100k runs) | Match? |
|---|---|---|---|
| $P(X \ge 8)$, guessing quiz | $\dfrac{56}{1024} \approx 0.0547$ | $\approx 0.055$ | within $\sim 0.001$ |
| $P(N = 4)$, shuttle arrivals | $e^{-4}4^4/4! \approx 0.195$ | $\approx 0.195$ | within $\sim 0.002$ |
| binomial pmf, all scores | $\binom{10}{x}(0.5)^{10}$ | empirical $\approx$ exact | differences scatter around $0$ |
| Poisson mean & variance | both $=\lambda = 4$ | sample mean $\approx 4$, var $\approx 4$ | both near $4$ |

The exact figures above are the ones derived by hand in the Week 9 notes; treat them as the answer key the
simulation is being checked against, not numbers the code invents. When you run the chunks, your estimates
will not be *identical* to mine to every decimal even with the same seed and run count — different R
versions can produce slightly different streams — but they should be close enough that the agreement is
obvious. That closeness is the result: two independent routes, the formula and the simulation, arriving at
the same probability is strong evidence you have the model right.

Two follow-ups worth a minute. First, **change the seed** to `set.seed(2026)` and rerun: every estimate
shifts a little, but the formula values do not move — that contrast shows you which number is random (the
estimate) and which is fixed (the formula). Second, **shrink the runs** to `n_sim <- 1000` and watch the
estimates get noticeably noisier; bump to `n_sim <- 1000000` and watch them tighten. That is the law of
large numbers in miniature, the very thing Week 13 will make formal. The gap between simulation and formula
is not error in the formula — it is the price of estimating, and it is one you can pay down with more runs.

## AI use note

If you used an AI assistant on this lab — to explain an `rbinom` argument, to debug a `table()` call, to
suggest a cleaner way to count a tail — log it briefly. The honest, load-bearing column is **Verification**:
how you confirmed the output yourself rather than trusting it. In this lab you are lucky, because you have a
built-in check — the formula. Anything the AI helps you compute can be held against `dbinom`/`dpois` or the
known values $0.0547$ and $0.195$.

| Tool | Purpose | Verification |
|---|---|---|
| AI chat assistant | Explain what `size` vs `n` mean in `rbinom(n, size, prob)` | Re-read `?rbinom` in R; confirmed `size` is questions-per-quiz, `n` is number of quizzes; checked the output vector had length `n_sim` with values in $0\text{–}10$ |
| AI chat assistant | Suggest how to count $P(X \ge 8)$ from the `scores` vector | Verified `mean(scores >= 8)` against `sum(dbinom(8:10, 10, 0.5))`; both $\approx 0.055$ |
| AI chat assistant | Help read a `table()` that dropped a missing score | Confirmed the fix by wrapping in `factor(..., levels = 0:10)` and checking all eleven scores now appear |

If you used no AI help, write one line saying so — that is a complete and honest note. The goal is a record
you would be comfortable showing, where the verification step proves the answer is yours.

## See also

- [Week 9 — Common discrete models](../notes/week-09-common-discrete-models.qmd) — the companion note where
  $P(X \ge 8) \approx 0.0547$ and $P(N = 4) \approx 0.195$ are derived by hand.
- [Lab 2 — Monte Carlo basics](lab-02-monte-carlo-basics.qmd) — the first "estimate a probability by
  counting" lab, which `mean(condition)` here builds on.
- [Lab 13 — Law of large numbers and CLT](lab-13-law-of-large-numbers-and-clt.qmd) — why more runs shrink
  the gap between simulation and formula, made precise.
- [Distribution reference](../resources/distribution-reference.qmd) — the binomial and Poisson parameters,
  means, and variances on one page.
- [R · Quarto setup](../resources/r-quarto-setup.qmd) — getting R, RStudio/Posit Cloud, and Quarto running.

*The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.*
