---
title: "Lab 5 — Bayes by simulation"
subtitle: "Checking the screening-test posterior by simulating a population"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** Week 5 computed the screening-test posterior $P(D\mid +)\approx 0.162$ by algebra — Bayes'
> rule with the denominator built from total probability. This lab reaches the *same* number a completely
> different way: we build a synthetic population of a hundred thousand people, test all of them, and then
> simply **count** what fraction of the positives actually have the condition. If the two answers agree, you
> have seen Bayes' rule confirmed by a population rather than asserted by a formula. The companion note is
> [Week 5 — Bayes' rule & updating](../notes/week-05-bayes-rule-and-updating.qmd); read it first, because
> this lab assumes the prior–likelihood–posterior vocabulary from there.

## The idea

Bayes' rule can feel like a sleight of hand. You start with a likelihood you *can* measure —
$P(+\mid D)=0.95$, the chance a sick person tests positive — and out comes a posterior you *want* —
$P(D\mid +)\approx 0.162$, the chance a person who tested positive is actually sick. The two numbers are far
apart, the small one is the right one, and it is reasonable to wonder whether the algebra is hiding
something. It is not. The cleanest way to convince yourself is to stop manipulating symbols and instead build
a world where the proportions are literally true, then go look.

That is the whole idea of this lab. We will create a synthetic population in which exactly the Week-5 numbers
hold: about $2\%$ of people have the condition, $95\%$ of those who have it test positive, and $10\%$ of
those who do not have it test positive anyway. Then we set the algebra aside entirely and ask one empirical
question of the data — **of everyone who tested positive, what fraction actually had the condition?** That
fraction *is* the posterior $P(D\mid +)$, defined the most concrete way possible: a conditional probability
is a proportion within a subgroup. If the population is large enough, that counted proportion should land
right on the $0.162$ the note derived by hand.

The discipline that makes this honest is worth stating up front: **we never feed the answer in.** Nowhere in
the simulation do we type $0.162$, and nowhere do we type the denominator $0.117$. We supply only the three
inputs a Week-5 problem hands you — the prevalence, the sensitivity, and the false-positive rate — and let
the posterior *and* the marginal $P(+)$ fall out of the counting. If we had typed the answer in, recovering
it would prove nothing. Because we do not, recovering it is a genuine check.

Before any code, here is the whole pipeline in one picture: build the population, split it by disease status,
split each group again by test result at *its own* rate, then count inside the positive pool alone.

![**The lab's pipeline (synthetic).** A population of $N=100{,}000$ is split by disease status (the prior),
then split again by test result conditional on that status (the likelihoods); only the positive pool is
counted to recover the posterior, which is then checked against the exact Bayes
value.](../assets/figures/lab05_pipeline_anchor.png){#fig-lab05-pipeline fig-alt="A top-to-bottom flow diagram: a population of 100,000 box leads to a disease-status box, which leads to a test-result box, which branches into a positive pool and a negative pool. The positive pool leads to a count-the-posterior box, which points to a box comparing the result to the exact Bayes posterior of 0.162."}

Each of the three boxes in the middle is exactly one of the three steps below; the comparison box on the
right is the Verify section at the end. Keep this picture in mind as a map — it is easy to lose the thread
inside the R code.

## Goal

By the end of this lab you should be able to:

- **Simulate a labelled population** in base R: draw each person's true disease status from a Bernoulli
  trial at the prevalence rate, then draw each person's test result from a Bernoulli trial whose success
  probability *depends on* that person's status.
- **Recover a conditional probability by counting**, using subsetting — `mean(disease[test == 1])` reads
  literally as "the proportion who have the disease, among those who tested positive."
- **Watch the Bayes denominator appear on its own**: the overall positive rate `mean(test)` should sit near
  $0.117$ without our ever having computed it, because total probability is just bookkeeping over the two
  groups.
- **Compare the simulated posterior to the exact value** $0.162$, and explain the small gap as **Monte Carlo
  error** — sampling noise that shrinks as the population grows.
- **Articulate the base-rate surprise** in population terms: a good test, applied to a rare condition,
  produces a pool of positives that is mostly false positives.

This is a study lab: all data are **synthetic, seed 35003 set**, and the code is shown for you to read and
run yourself.

## Setup

You need nothing beyond base R — no packages. If you have not set up R, RStudio (or Posit Cloud), and
Quarto yet, the [R · Quarto setup](../resources/r-quarto-setup.qmd) page walks through it. Open a fresh
script or a Quarto document and keep these three facts from Week 5 in front of you, because they are the
*only* inputs the simulation is allowed to know:

| Quantity | Symbol | Value | Role in Bayes |
|---|---|---|---|
| Prevalence | $\pi = P(D)$ | $0.02$ | the **prior** |
| Sensitivity | $P(+\mid D)$ | $0.95$ | the **likelihood** if diseased |
| False-positive rate | $P(+\mid D^{c})$ | $0.10$ | the **likelihood** if healthy |

A note on the false-positive rate: Week 5 was handed *specificity* $P(-\mid D^{c}) = 0.90$, and took the
complement to get $P(+\mid D^{c}) = 1 - 0.90 = 0.10$. We will do that one complement explicitly in the code
so the connection to the note is visible, then use $0.10$ as the healthy-person positive rate.

The two targets we are trying to *recover* — and which we therefore do **not** type as inputs — are the
marginal $P(+) \approx 0.117$ and the posterior $P(D\mid +) \approx 0.162$. Keep them on a sticky note as
answers to check against, not as numbers to enter.

The population size is a modelling choice. We use $N = 100{,}000$. That is large enough that the counted
proportions are stable to about two decimal places, yet small enough to run in a blink. Setting
`set.seed(35003)` makes the whole thing reproducible: you and a classmate running the same code get the same
"random" population, so you can compare results exactly.

```{r}
#| label: setup-inputs
#| eval: false

set.seed(35003)

# Population size and the THREE allowed inputs (the Week-5 givens).
N           <- 100000
prevalence  <- 0.02   # pi  = P(D)        -- the prior
sensitivity <- 0.95   #       P(+ | D)    -- likelihood if diseased
specificity <- 0.90   #       P(- | not D)
fpr         <- 1 - specificity   # P(+ | not D) = 0.10, the false-positive likelihood

# We deliberately do NOT define the posterior or the denominator here.
# Both must FALL OUT of the counting below, or the check proves nothing.
```

## Steps

The plan has three moves: build true disease status, build the test result *conditional on* that status, and
then count the posterior. Read each chunk before you run it — the point of the lab is understanding the
construction, not just seeing the final number.

### Step 1 — Simulate true disease status

First give every one of the $N$ people a true, hidden disease status. Each person independently "has the
condition" with probability equal to the prevalence, $\pi = 0.02$. That is exactly a Bernoulli trial, and
`rbinom(N, size = 1, prob = prevalence)` draws $N$ of them at once: it returns a vector of $N$ values, each a
$1$ ("has the condition") or a $0$ ("does not"). Think of `disease` as the column of ground truth that, in
real life, you never get to see directly — the test is your only window onto it.

```{r}
#| label: step1-disease-status
#| eval: false

set.seed(35003)

N          <- 100000
prevalence <- 0.02

# Each person: 1 = has the condition, 0 = does not. Bernoulli(prevalence).
disease <- rbinom(N, size = 1, prob = prevalence)

# How many true cases did this synthetic world produce?
sum(disease)         # roughly prevalence * N  ~ 0.02 * 100000 = 2000
mean(disease)        # the realized prevalence; should be close to 0.02
```

Two sanity checks are built in. `sum(disease)` counts the true cases — expect something near
$0.02 \times 100{,}000 = 2{,}000$, though not exactly, because the count is itself random. `mean(disease)`
reports the *realized* prevalence in this particular population; it should sit close to $0.02$ but rarely
equal it on the nose. That small wobble is your first sight of Monte Carlo error: even the input rate is only
*approximately* reproduced in any finite sample. Hold that thought — it explains the gap we will see at the
end.

The figure below is what this chunk's two sanity checks actually produced in one run: a population split into
a small diseased group and a large healthy group.

![**Step 1's population split (synthetic).** Running the chunk above once produced $1{,}993$ diseased people
($1.99\%$) against $98{,}007$ healthy people ($98.01\%$) — close to, but not exactly, the input $0.02$; that
small gap is Monte Carlo error, visible before the test even
enters.](../assets/figures/lab05_disease_split_bars.png){#fig-lab05-disease-split fig-alt="A bar chart with two bars: a very short bar labeled disease equals 1 at 1,993 people, and a much taller bar labeled disease equals 0 at 98,007 people."}

Notice how short the "Disease = 1" bar is next to "Disease = 0" — that size gap between the two groups is the
whole reason the base-rate surprise happens, and Step 3 will make its consequence concrete.

### Step 2 — Simulate the test, conditional on status

Now test everyone. The crucial point — and the place where Bayes lives — is that **the test's behaviour
depends on the person's true status.** A diseased person tests positive with probability $0.95$
(the sensitivity); a healthy person tests positive with probability $0.10$ (the false-positive rate). So each
person does not share one positive-probability; each carries their *own*, set by their `disease` value.

We build a per-person probability vector with `ifelse`: wherever `disease == 1`, use `sensitivity`; wherever
`disease == 0`, use `fpr`. Then a single `rbinom` call draws each person's test result from *their own*
Bernoulli trial. Passing a whole vector to the `prob` argument is the idiom that makes this conditional
simulation a one-liner.

```{r}
#| label: step2-test-given-status
#| eval: false

set.seed(35003)

# (Re-run Step 1's draws so this chunk stands alone.)
N           <- 100000
prevalence  <- 0.02
sensitivity <- 0.95
specificity <- 0.90
fpr         <- 1 - specificity        # 0.10

disease <- rbinom(N, size = 1, prob = prevalence)

# Each person's OWN positive-probability, chosen by their true status:
#   diseased  -> sensitivity (0.95)
#   healthy   -> fpr         (0.10)
p_positive <- ifelse(disease == 1, sensitivity, fpr)

# Draw each test result from that person-specific Bernoulli trial.
test <- rbinom(N, size = 1, prob = p_positive)

# The OVERALL positive rate -- the Bayes denominator P(+) -- falls out here.
mean(test)           # expect ~ 0.117, and we never typed 0.117 anywhere
```

Look hard at `mean(test)`. We never computed the denominator $P(+) = 0.95(0.02) + 0.10(0.98) = 0.117$; we
only told the simulation how each *group* behaves. Yet the overall positive rate lands near $0.117$ all the
same. That is the **law of total probability happening physically**: the population splits into a true-case
group and a healthy group, each contributes positives at its own rate, and the marginal is just the
size-weighted blend of the two. The formula in the note and this counted average are two descriptions of one
fact.

One structural caution worth noting now. The diseased group is *small* — only about $2{,}000$ people out of
$100{,}000$. The vast healthy group, $98{,}000$ strong, throws off false positives at a $10\%$ clip, which is
roughly $9{,}800$ of them. So the positive pool is dominated by healthy people. That imbalance is precisely
the base-rate surprise, and Step 3 will read its consequence straight off the data.

Splitting the same run into all four (disease, test) combinations makes the imbalance concrete:

![**Each group tested at its own rate (synthetic).** Of the diseased group, $1{,}876$ tested positive and
$117$ tested negative (a $95\%$-ish hit rate). Of the far larger healthy group, $9{,}849$ still tested positive
(the $10\%$ false-positive rate applied to $98{,}007$ people) against $88{,}158$ negatives. The overall
`mean(test)` this run gives is $0.1172$ — close to the exact marginal $P(+)=0.117$, though $0.117$ was never
typed in anywhere.](../assets/figures/lab05_four_group_population.png){#fig-lab05-four-group fig-alt="A grouped bar chart with two clusters. The diseased-group cluster shows 1,876 tested positive and 117 tested negative. The healthy-group cluster shows 9,849 tested positive and 88,158 tested negative."}

Compare the two "tested positive" bars: $1{,}876$ from the tiny diseased group versus $9{,}849$ from the
enormous healthy group. Positives are already outnumbered by false alarms before you compute anything else —
that is exactly what Step 3 counts directly.

### Step 3 — Recover the posterior by counting

Now the payoff. The posterior $P(D\mid +)$ is, by definition, a conditional probability: the probability of
disease *among the subgroup that tested positive*. In data terms that is a proportion within a subset — so we
subset. The expression `disease[test == 1]` keeps the true-status values for **only** the people who tested
positive, and `mean(...)` of that subset is the fraction of them who actually have the condition. That single
number is the simulated posterior.

```{r}
#| label: step3-posterior-by-counting
#| eval: false

set.seed(35003)

# (Stand-alone: rebuild the population.)
N           <- 100000
prevalence  <- 0.02
sensitivity <- 0.95
specificity <- 0.90
fpr         <- 1 - specificity

disease    <- rbinom(N, size = 1, prob = prevalence)
p_positive <- ifelse(disease == 1, sensitivity, fpr)
test       <- rbinom(N, size = 1, prob = p_positive)

# Among everyone who tested positive, what fraction truly has the condition?
# disease[test == 1]  : true-status values for the positive subgroup only
# mean(...)           : the proportion of that subgroup with disease == 1
posterior_sim <- mean(disease[test == 1])
posterior_sim        # expect ~ 0.162

# It helps to see the counts behind the proportion:
n_positive      <- sum(test == 1)                 # size of the positive pool
n_true_positive <- sum(disease == 1 & test == 1)  # truly sick AND positive
n_positive
n_true_positive
n_true_positive / n_positive   # same as posterior_sim, written as a ratio
```

The last line spells out what `mean(disease[test == 1])` is doing: it is `n_true_positive / n_positive`, the
share of the positive pool that is genuinely sick. Run it and you should see roughly $0.16$ — about one in
six positives is a true case, and the other five in six are false alarms. We arrived there with no Bayes
formula in sight; the population did the bookkeeping for us, and `mean` read off the answer.

A small detail keeps this honest: at no point did we type the posterior or the denominator. The inputs were
only `prevalence`, `sensitivity`, and `fpr`. Everything else — the marginal $0.117$ in Step 2, the posterior
$0.162$ here — emerged from counting. That is what makes the agreement in the next section meaningful rather
than circular.

The picture below is the payoff of the whole lab, drawn directly from this run's counts: **inside the positive
pool only**, who is actually sick?

![**Inside the positive pool (synthetic).** Of the $11{,}725$ people who tested positive in this run, only
$1{,}876$ ($16.0\%$) truly have the condition; the other $9{,}849$ ($84.0\%$) are healthy false positives. A
positive result on this good test still leaves most people
healthy.](../assets/figures/lab05_positive_pool_breakdown.png){#fig-lab05-positive-pool fig-alt="A horizontal bar chart with two bars restricted to the positive-testing subgroup: a short bar labeled truly diseased, true positive at 1,876 people (16.0 percent of positives), and a much longer bar labeled actually healthy, false positive at 9,849 people (84.0 percent of positives)."}

This is the base-rate surprise, seen as two stacks of people rather than asserted in words: **most of the
positive pool is false alarms**, not because the test is bad, but because the disease is rare. The short bar
in this figure, divided by the pool's total, is exactly `posterior_sim`.

## Verify

Now hold the simulated posterior up against the exact value the note derived by hand. Compute the Bayes
number directly from the *same three inputs* and compare:

```{r}
#| label: verify-against-exact
#| eval: false

# Exact Bayes from the three inputs (mirrors the Week-5 derivation):
p_marginal_exact  <- sensitivity * prevalence + fpr * (1 - prevalence)   # P(+)
posterior_exact   <- (sensitivity * prevalence) / p_marginal_exact       # P(D | +)

p_marginal_exact   # 0.117
posterior_exact    # ~ 0.1623932 ... rounds to 0.162

# Side-by-side with the simulated values from the steps above:
c(simulated_marginal  = mean(test),
  exact_marginal      = p_marginal_exact,
  simulated_posterior = posterior_sim,
  exact_posterior     = posterior_exact)
```

This chunk's output, from an actual run, is shown side by side with the exact algebra below:

![**Counted vs. exact, side by side (synthetic).** Simulated (counted) values from this run —
marginal $0.1172$, posterior $0.1600$ — plotted against the exact algebraic values — marginal $0.117$,
posterior $0.1624$. The bars are nearly the same
height.](../assets/figures/lab05_sim_vs_exact_comparison.png){#fig-lab05-verify-compare fig-alt="A grouped bar chart with two pairs of bars. The marginal P plus pair shows simulated 0.1172 next to exact 0.1170. The posterior P D given plus pair shows simulated 0.1600 next to exact 0.1624."}

You should find the simulated and exact values agree to about two decimal places — the marginal near
$0.117$, the posterior near $0.162$. They will not match *exactly*, and that is correct and expected. The
simulation estimates the posterior from a finite population, so it inherits sampling noise; this is **Monte
Carlo error**. Two things govern its size. First, the *whole* population is large ($N = 100{,}000$), which
keeps the estimate stable. Second — and this is the real bottleneck here — the posterior is a proportion
computed *within the positive subgroup*, and only the truly-sick-and-positive people (around $1{,}900$ of
them) carry the signal. A proportion based on a couple thousand cases still jitters in the third decimal. If
you want a sharper estimate, raise $N$ (say to a million) and rerun; the counted posterior will tighten
toward $0.162$. If you want to *see* the noise, change the seed a few times and watch the third decimal
move while the first two hold steady.

Both of those knobs are visible in the same picture:

![**Two ways to shrink Monte Carlo error (synthetic).** The black line tracks the counted posterior as $N$
grows from $1{,}000$ to $1{,}000{,}000$, closing in on the exact $0.1624$ (dashed line). The open circles hold
$N=100{,}000$ fixed — this lab's own population size — and simply change the seed six times; the estimate
jitters around the same
target.](../assets/figures/lab05_convergence_by_n.png){#fig-lab05-convergence fig-alt="A line plot with population size N on a log scale from 1,000 to 1,000,000 on the x-axis and the counted posterior on the y-axis. A line of filled points rises from about 0.131 at N equals 1,000 toward about 0.162 at N equals 1,000,000, approaching a dashed horizontal line at the exact value 0.1624. A cluster of six open circles at N equals 100,000 scatters above and below the line, showing seed-to-seed variation at fixed N."}

Raising $N$ moves you left-to-right along the black line; changing the seed at fixed $N$ moves you up and
down within the cluster of open circles. Both shrink the *gap* to the dashed exact line, but only raising $N$
shrinks it systematically — a new seed can land you closer or farther away by chance.

Finally, read the base-rate surprise off your own counts. The positive pool had roughly $11{,}700$ people in
it, but only about $1{,}900$ were truly sick — so a positive result, on this genuinely good test, still
leaves a person far more likely healthy than not. The reason is structural and the simulation makes it
tangible: a rare condition means the healthy group, even at a modest $10\%$ false-positive rate, simply
outnumbers the true cases in the positive pool. The posterior is small not because the test is bad but
because the prior was small — the same lesson Week 5 drew from algebra, now visible as two stacks of people.

## AI use note

If you used an AI assistant while working through this lab — to explain `ifelse` on a vector, to debug a
subsetting error, or to reword a confusing message — record it honestly. The verification column is the one
that matters: it is where you say how you confirmed the help was right, because an AI can produce confident
nonsense and the responsibility for the result is yours.

| Tool | Purpose | Verification |
|---|---|---|
| AI assistant (general) | Explain why `prob = p_positive` accepts a *vector* in `rbinom` | Re-read `?rbinom`; confirmed `prob` is vectorized by testing a tiny 4-person vector by hand |
| AI assistant (general) | Debug a wrong posterior caused by `disease[test = 1]` (one `=`) | Reproduced the error, fixed to `test == 1`, and re-checked that `mean(disease[test == 1])` returned ~0.16 |
| AI assistant (general) | Reword the Monte-Carlo-error explanation in my own notes | Cross-checked the claim against the Verify section and rewrote it in my own words, citing $N$ and the subgroup size |

Keep the table short and specific. "I used AI to do the lab" is not a record; "I used AI to explain X, and I
verified it by Y" is.

## See also

- [Week 5 — Bayes' rule & updating](../notes/week-05-bayes-rule-and-updating.qmd) — the companion note: the
  algebraic derivation of the same posterior this lab recovers by counting.
- [Lab 2 — Monte Carlo basics](lab-02-monte-carlo-basics.qmd) — the estimate-a-probability-by-counting habit
  this lab leans on, in its simplest form.
- [Week 4 — Independence & information](../notes/week-04-independence-and-information.qmd) — why the test and
  the disease status are *not* independent, which is exactly what makes the conditional simulation work.
- [R · Quarto setup](../resources/r-quarto-setup.qmd) — getting R, RStudio/Posit Cloud, and Quarto running so
  you can execute these chunks yourself.
- [Notation glossary](../resources/notation-glossary.qmd) — the binding symbols, including prior, likelihood,
  evidence, and posterior.

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
