---
title: "Lab 2 — Monte Carlo basics"
subtitle: "Estimating a probability by repeated trials and counting"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab turns a probability you already trust from the notes into something you can *watch*
> happen. In [Week 2 — Sample spaces, events & rules](../notes/week-02-sample-spaces-events-rules.qmd) we
> said Maya's morning shuttle is on time on a fraction $0.81$ of days, and that this number means a long-run
> frequency. Here you will generate a long run of synthetic mornings, count the on-time ones, and see the
> running fraction settle toward $0.81$ — the idea behind every Monte Carlo estimate in the rest of the
> course. All data are **synthetic; seed `35003` set**.

## The idea

A probability is, at heart, a statement about the long run. When the notes say $P(\text{on time}) = 0.81$,
they are claiming that if Maya could relive an enormous number of statistically identical mornings, the
shuttle would be on time on close to $81\%$ of them. That is a claim about frequency, and frequency is
something a computer can manufacture on demand. We do not have to wait years for real mornings to accumulate;
we can ask R to produce thousands of synthetic ones in a fraction of a second, each one on time or late with
the right chance, and then simply count.

That is the whole of **Monte Carlo estimation**: to approximate a probability, simulate the random situation
many times and report the fraction of simulations in which the event of interest occurred. The name is a nod
to games of chance, and the spirit is exactly that — play the game enough times and the scoreboard tells you
the odds. The estimate is never exactly right on any finite run, but it has a reassuring property: the more
trials you run, the closer it tends to land, and the *typical* size of its error shrinks in a predictable
way. By the end of this lab you will have built the estimate, watched it converge, and seen why running more
trials buys you accuracy — but at a slowing rate.

Three ideas carry the whole lab, and it is worth naming them before any code appears:

- **A probability is a long-run frequency.** The on-time fraction over many mornings *is* what
  $P(\text{on time})$ is approximating. Simulation makes this definition tangible rather than abstract.
- **Monte Carlo error shrinks like $1/\sqrt{n}$.** Quadruple the number of trials and you roughly halve the
  typical distance between your estimate and the truth. Accuracy is buyable, but it gets expensive.
- **Reproducibility comes from the seed.** A simulation uses a pseudo-random number stream. Fix its starting
  point with `set.seed(35003)` and anyone who runs your code gets the identical "random" mornings — the same
  numbers, the same plot, the same conclusion. That is what makes a simulation a *result* and not a one-off
  anecdote.

The four numbered Steps below are not four unrelated exercises — they are one five-part recipe, always
performed in the same order.

![**The Monte Carlo recipe this lab walks (synthetic).** Model one morning as a weighted coin, simulate many
mornings, count and divide to get a point estimate, track that estimate as a running fraction, then compare it
to the theoretical value.](../assets/figures/lab02_mc_pipeline.png){#fig-lab02-pipeline fig-alt="A left-to-right flow diagram of five boxes connected by arrows: model one morning as a weighted coin with p = 0.81; simulate n = 10000 synthetic mornings; count and divide, p_hat = mean(trials); track the running fraction as n grows; compare p_hat to theory, 0.81."}

*Non-visual equivalent:* the recipe in words is (1) model a morning as a weighted coin with $P(\text{on
time})=0.81$, (2) simulate `n_trials` such mornings, (3) estimate by counting (`p_hat = mean(trials)`), (4)
track the running fraction as more mornings accumulate, (5) compare that running fraction to the theoretical
$0.81$. Steps 1–4 below build exactly these five pieces, in this order.

## Goal

By the end of this lab you will be able to:

1. **Model a single random trial** — one of Maya's mornings — as a weighted coin flip that lands "on time"
   with probability $0.81$ and "late" with probability $0.19$, using `rbinom()` with `size = 1`.
2. **Estimate a probability by counting** — generate many such trials and take their mean, so that the
   sample fraction of on-time mornings becomes your Monte Carlo estimate of $P(\text{on time})$.
3. **Watch the estimate converge** — compute the *running* relative frequency as trials accumulate and see
   it tighten around the theoretical value $0.81$.
4. **Compare simulation to theory** — read the gap between your estimate and $0.81$ as Monte Carlo error,
   and explain why that gap tends to shrink like $1/\sqrt{n}$ rather than disappearing all at once.

Everything here is base R. The code is **shown for study** and is marked so that it does not execute on this
site; you run it yourself in your own R session, where the fixed seed makes the output reproducible.

## Setup

You need nothing beyond a working R installation and, ideally, the project setup described on the
[R · Quarto setup](../resources/r-quarto-setup.qmd) page. No add-on packages are required — `rbinom()`,
`mean()`, `cumsum()`, and `plot()` are all part of base R.

The situation we are modeling is the recurring one from the notes. Maya catches a campus shuttle each
weekday morning. On any given morning the shuttle is either **on time** or **late**, and from the Week 2 and
Week 3 development we take the long-run on-time chance as a fixed, known number:

$$
P(\text{on time}) = 0.81, \qquad P(\text{late}) = 1 - 0.81 = 0.19 .
$$

We will treat each morning as an independent trial with this fixed success probability — a **weighted coin
flip** where "heads" means on time. That assumption (independent, identically weighted mornings) is exactly
what makes the long-run-frequency picture clean, and it is the assumption we are choosing to simulate. The
numbers are **synthetic; seed `35003` set** throughout.

Two small conventions make the code below easy to read:

- We encode each trial as a number: **`1` = on time**, **`0` = late**. With that coding, the *count* of
  on-time mornings is just the sum of the trials, and the *fraction* on time is their mean.
- We set the seed **once**, at the very top of the script, before any random numbers are drawn. Re-running
  from that line reproduces the entire lab.

```{r}
#| label: setup-constants
#| eval: false

# Lab 2 — Monte Carlo basics. Synthetic data; seed set for reproducibility.
set.seed(35003)

p_on_time <- 0.81   # theoretical P(on time) from the Week 2/3 notes
n_trials  <- 10000  # number of synthetic mornings to simulate
```

Read the two constants as the two things a Monte Carlo study always needs: the *truth we are trying to
recover* (`p_on_time`, here known, because this is a teaching example) and the *amount of effort we will
spend* (`n_trials`). In a real application you would not know `p_on_time` — that is what you would be
estimating. We keep it on hand here precisely so we can check the estimate against it at the end.

## Steps

### Step 1 — Simulate the trials

A single morning is a weighted coin flip. The base-R tool for "flip a coin that lands `1` with probability
`p`" is `rbinom(n, size = 1, prob = p)`: it returns `n` independent draws, each a `0` or a `1`, where `1`
appears with probability `prob`. Setting `size = 1` is what makes each draw a single trial (a *Bernoulli*
trial) rather than a count over several flips. Asking for `n_trials` of them in one call gives us the whole
run of synthetic mornings at once.

```{r}
#| label: simulate-trials
#| eval: false

set.seed(35003)

# Each entry is one morning: 1 = on time, 0 = late.
trials <- rbinom(n = n_trials, size = 1, prob = p_on_time)

# A first look at what we made.
length(trials)      # how many mornings: 10000
head(trials, 20)    # the first 20 mornings as 0s and 1s
sum(trials)         # how many were on time, in total
```

The vector `trials` is the simulated long run: ten thousand mornings, each a `0` or a `1`. There is nothing
mysterious in it — it is just a list of coin flips weighted toward "on time." `sum(trials)` counts the
on-time mornings, and because on-time is coded `1`, that sum *is* the count of the event we care about.
Notice that we call `set.seed(35003)` again here. That is deliberate: it lets this step stand on its own and
reproduce the exact same `trials` whether or not you ran the setup chunk first.

The chunk above is shown but not executed on this page. Because the seed is fixed, running it yourself
produces one exact, reproducible sequence — the figure below shows precisely what that sequence looks like.

![**The first 20 synthetic mornings, and the full-run total (synthetic; seed `35003`).** The strip reproduces
`head(trials, 20)` exactly; the caption line reproduces `sum(trials)` over all `n_trials = 10000` mornings.](../assets/figures/lab02_first20_trials_strip.png){#fig-lab02-first20 fig-alt="A row of 20 circles labeled 0 or 1, colored by outcome: on time (filled, darker) or late (open, lighter). Reading left to right the sequence is 0,0,1,1,0,0,0,0,1,1,1,1,1,1,0,0,1,1,1,0. A caption below states that out of all 10000 simulated mornings, 8121 were on time, giving p_hat = 0.8121."}

*Non-visual equivalent:* `head(trials, 20)` is `0 0 1 1 0 0 0 0 1 1 1 1 1 1 0 0 1 1 1 0`; `sum(trials)` over
all 10,000 mornings is $8121$ on-time mornings.

### Step 2 — Estimate the probability by counting

The Monte Carlo estimate of $P(\text{on time})$ is the fraction of mornings that came up on time:

$$
\widehat{P}(\text{on time}) = \frac{\text{number of on-time mornings}}{\text{number of mornings}}
= \frac{1}{n}\sum_{i=1}^{n} \text{trial}_i .
$$

With our `0`/`1` coding, that fraction is exactly the **mean** of the `trials` vector — summing `1`s and
dividing by the count is what `mean()` does. So the estimate is a single, almost anticlimactic line of code.

```{r}
#| label: estimate-by-counting
#| eval: false

# The Monte Carlo estimate is just the sample fraction of on-time mornings.
n_on_time <- sum(trials)          # count of the event
p_hat     <- mean(trials)         # equivalently n_on_time / n_trials

n_on_time
p_hat                             # our estimate of P(on time)
```

`p_hat` is the lab's headline number: the simulation's verdict on how often the shuttle is on time, read off
purely by counting and dividing. It will be *close* to $0.81$ but, on any finite run, not exactly equal to
it. That small gap is not a mistake — it is **Monte Carlo error**, the price of estimating a probability from
a finite sample instead of computing it exactly. The next step shows where that gap comes from and how it
behaves as the run grows.

![**One Monte Carlo estimate, next to the theoretical value (synthetic; seed `35003`).** The point marks
`p_hat`; the dashed line marks the theoretical `p_on_time = 0.81` this run is trying to
recover.](../assets/figures/lab02_point_estimate_bar.png){#fig-lab02-estimate fig-alt="A number line from 0.0 to 1.0. A dashed vertical line marks theory at 0.81. A solid point, labeled p_hat = 0.8121, sits just to the right of the dashed line."}

*Non-visual equivalent:* `n_on_time = 8121`, `p_hat = 0.8121`, a gap from theory of about $0.0021$ — a single
estimate landing close to, but not exactly on, $0.81$.

### Step 3 — Watch the running relative frequency converge

A single estimate hides the most instructive part of the story: *how* the fraction settles down as mornings
accumulate. To see it, we compute the **running relative frequency** — after morning 1, after morning 2,
after morning 3, and so on — and look at the whole trajectory. The running count of on-time mornings up to
each point is a cumulative sum (`cumsum`), and dividing by the morning index turns each running count into a
running fraction.

```{r}
#| label: running-frequency
#| eval: false

# Running fraction of on-time mornings after 1, 2, 3, ..., n trials.
running_count <- cumsum(trials)           # on-time mornings so far
index         <- seq_along(trials)        # 1, 2, 3, ..., n_trials
running_freq  <- running_count / index    # the estimate after each morning

# The estimate is volatile early and steadies as n grows.
running_freq[c(10, 100, 1000, 10000)]     # snapshots at four sample sizes
```

Reading the four snapshots from left to right tells the convergence story in one line: at $n = 10$ the
estimate can be well off, at $n = 100$ it is closer, and by $n = 1000$ and $n = 10000$ it is hugging $0.81$.
Plotting the whole path makes the behavior unmistakable — early wild swings that calm into a near-flat line
near the true value.

![**Four snapshots of the running fraction (synthetic; seed `35003`).** The same `running_freq` vector, read
off at $n = 10, 100, 1000, 10000$ — coarse frames from the full movie the next figure
plays.](../assets/figures/lab02_snapshot_dotplot.png){#fig-lab02-snapshots fig-alt="A dot plot on a log-scaled horizontal axis of n = 10, 100, 1000, 10000. The running fraction at each n is 0.400, 0.840, 0.807, 0.812. A dashed horizontal line marks theory at 0.81; the last two points sit almost on it."}

*Non-visual equivalent:* `running_freq[c(10, 100, 1000, 10000)]` is $0.400,\ 0.840,\ 0.807,\ 0.812$ — noisy at
$n=10$, then progressively closer to $0.81$.

```{r}
#| label: plot-convergence
#| eval: false

plot(index, running_freq, type = "l",
     xlab = "number of mornings simulated (n)",
     ylab = "running fraction on time",
     main = "Monte Carlo estimate settling toward 0.81")
abline(h = p_on_time, lty = 2)            # the theoretical value 0.81
```

The dashed horizontal line marks the truth, $0.81$. The solid curve is your estimate as a function of how
many mornings you have counted. The shape is the whole lesson: the curve does not march smoothly to the line,
it *rattles* toward it, with the rattling growing quieter as $n$ grows. The narrowing of those swings is the
$1/\sqrt{n}$ behavior made visible — to halve the typical width of the wiggle you must run roughly *four
times* as many trials.

![**The convergence plot the chunk above would produce (synthetic; seed `35003`).** This is exactly the
`plot(index, running_freq, ...)` call, run once and captured: wide early swings that visibly narrow and settle
against the dashed $0.81$ line.](../assets/figures/lab02_convergence_plot.png){#fig-lab02-convergence fig-alt="A line plot of the running fraction on time against the number of mornings simulated, from 1 to 10000. The line starts volatile, swinging widely in the first few hundred mornings, then narrows and settles just around a dashed horizontal line at 0.81."}

*Non-visual equivalent:* the running fraction swings between roughly $0.4$ and $0.85$ within the first few
hundred mornings, then settles within about $\pm 0.01$ of $0.81$ by $n \approx 2000$ and stays there.

### Step 4 — Compare to the theoretical 0.81

The last step makes the comparison explicit. We report the estimate, the truth, and the gap between them —
the realized Monte Carlo error on this run — and put it next to a back-of-envelope sense of how big that gap
should be.

```{r}
#| label: compare-to-theory
#| eval: false

# Estimate vs. theory, and the realized error on this run.
abs_error <- abs(p_hat - p_on_time)

c(estimate    = p_hat,
  theoretical = p_on_time,
  abs_error   = abs_error)

# A rough yardstick for the typical error at this n:
# sqrt(p(1-p)/n) is the standard deviation of the estimate.
typical_error <- sqrt(p_on_time * (1 - p_on_time) / n_trials)
typical_error
```

Two numbers are worth dwelling on. The first is `abs_error`: how far *this particular* simulation landed
from $0.81$. The second is `typical_error`, the quantity $\sqrt{p(1-p)/n}$, which is the standard deviation
of the Monte Carlo estimate — a rule-of-thumb for how far off a run of this size *usually* is. You should
find `abs_error` is comfortably within a small multiple of `typical_error`; that is the sense in which the
estimate is "as accurate as it should be" for the effort spent. And because $n$ sits *under a square root* in
`typical_error`, the message is the same one the convergence plot drew: accuracy improves with more trials,
but only as fast as $1/\sqrt{n}$.

![**Is the realized error the right size (synthetic; seed `35003`)?** `abs_error` placed against `1x`, `2x`,
and `3x` of `typical_error` — the yardstick for deciding whether a gap from theory is ordinary Monte Carlo
noise or a sign something is wrong.](../assets/figures/lab02_error_comparison_dotplot.png){#fig-lab02-error fig-alt="A number line with tick marks at 1, 2, and 3 times typical_error (0.0039, 0.0078, 0.0118). A point marking abs_error = 0.0021 sits to the left of all three ticks, comfortably inside 1x typical_error."}

*Non-visual equivalent:* `typical_error` $\approx 0.0039$; this run's `abs_error` $\approx 0.0021$, which is
about $0.5$ times `typical_error` — comfortably inside the "small multiple" you should expect, not zero and
not many multiples too large.

**A common misreading to avoid:** neither an error of exactly zero nor a very large error is the reassuring
outcome. Zero would be a suspiciously perfect finite-sample run; several multiples of `typical_error` would
suggest a bug (a mistyped `prob`, or reusing an already-consumed random stream) rather than ordinary
Monte Carlo noise. "Comfortably within a small multiple" is the target, not "as close as possible."

## Verify

A simulation result is trustworthy only if it agrees with something you already know by other means. Here the
"other means" is the theory from the notes, and there are three checks worth making before you believe your
own output:

1. **Does the estimate match the theory?** Compare `p_hat` to $0.81$. With ten thousand trials they should
   agree to roughly two decimal places. If `p_hat` is wildly off — say $0.5$ or $0.95$ — suspect the
   `prob` argument, not the universe: a mistyped probability is the usual culprit.
2. **Does the running frequency settle, not wander?** The convergence plot should show large early swings
   that visibly damp down and hug the dashed line near $0.81$ as $n$ grows. A curve that *keeps* swinging
   at large $n$, or drifts toward a different level, means the trials are not what you think they are.
3. **Is the realized error the right size?** Confirm that `abs_error` is within a small multiple (say, two
   or three times) of `typical_error` $= \sqrt{p(1-p)/n}$. An error many times larger than that is a red
   flag; an error of *exactly* zero on a finite run would be suspicious in the other direction.

**A portability check.** Nothing about the four steps above is special to $0.81$. The Week 2 sample space also
gives us $P(\text{rain}) = 0.30$; swap `prob = 0.81` for `prob = 0.30` in `simulate-trials`, re-run the
identical four steps with the same `set.seed(35003)`, and you should see a running-frequency curve that
rattles early and settles — not near $0.81$ this time, but near $0.30$. The figure below shows exactly that
run, so you can check your own re-run against it before you trust the recipe on a probability of your own
choosing.

![**The same recipe, aimed at a different locked probability (synthetic; seed `35003`).** Everything is
identical to the Step 1–4 code except `prob = 0.30` in place of `prob = 0.81` — the shuttle's chance of rain,
not of being on time.](../assets/figures/lab02_transfer_rain_convergence.png){#fig-lab02-transfer fig-alt="A line plot of the running fraction of rainy mornings against the number of mornings simulated, from 1 to 10000. The line starts volatile near 1.0, drops and swings, then narrows and settles just around a dashed horizontal line at 0.30."}

*Non-visual equivalent:* with `prob = 0.30` and the same `n_trials = 10000` and seed, the final running
fraction is about $0.298$ — close to the locked $P(\text{rain}) = 0.30$, by the identical count-and-divide
logic used throughout this lab.

One more reproducibility check ties it all together: run the whole script a second time from `set.seed(35003)`
and confirm you get the **identical** `p_hat`, the identical plot, and the identical `abs_error`. A
Monte Carlo result that changes every time you run it is not yet a result — fixing the seed is what makes it
one. If a classmate runs your code on their machine and reports a different number, the first thing to check
is whether the seed line ran before any random draw.

## AI use note

If you use an AI assistant while working through this lab, record it briefly and honestly. The load-bearing
column is **Verification** — what *you* did to confirm the AI's output was correct, because you, not the
tool, are responsible for the result. Original work is expected; the assistant is a study aid, not an author.

| Tool | Purpose | Verification |
|---|---|---|
| AI chat assistant | Explain what `rbinom(n, size = 1, prob = p)` returns and how `size = 1` makes a single Bernoulli trial | Read the base-R `?rbinom` help page myself; ran a tiny call with `prob = 0.5` and checked the mean came out near $0.5$ |
| AI chat assistant | Suggest how to compute a running relative frequency in base R | Confirmed `cumsum(trials) / seq_along(trials)` against a hand-checked first three entries; checked the final entry equals `mean(trials)` |
| AI coding helper | Draft the convergence-plot call | Compared the plotted limit to the theoretical $0.81$ via the dashed `abline(h = ...)`; re-ran from `set.seed(35003)` to confirm the figure was reproducible |

## See also

- [Week 2 — Sample spaces, events & rules](../notes/week-02-sample-spaces-events-rules.qmd) — the companion
  note, where $P(\text{on time}) = 0.81$ is defined and read as a long-run frequency.
- [Lab 5 — Bayes by simulation](lab-05-bayes-by-simulation.qmd) — the same count-and-divide idea taken one
  step further, to recover a conditional (posterior) probability by simulating a whole population.
- [R · Quarto setup](../resources/r-quarto-setup.qmd) — how to get R, RStudio/Posit Cloud, and Quarto
  running so you can execute these chunks yourself.
- [Distribution reference](../resources/distribution-reference.qmd) — a quick reference for the base-R
  random-draw functions, including `rbinom` and the Bernoulli/Binomial parameterization used here.

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
