---
title: "Lab 13 — Law of large numbers and CLT"
subtitle: "Watching averages settle and sums turn bell-shaped"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 13 — Sums, simulation & limit behavior](../notes/week-13-sums-simulation-limit-behavior.qmd).
> The note states the two big limit theorems in words and symbols; here you *watch* them happen. You will
> build one long stream of simulated commute times and see its running average drift toward a fixed number
> (the law of large numbers), and you will build a whole crowd of sample means and see their histogram take
> on a bell shape that gets narrower as the sample size grows (the central limit theorem). Nothing here is
> graded — it is a guided rehearsal you can run in your own R session.

## The idea

All term the recurring "commuter's morning" world has given us numbers we treated as *fixed truths about the
process*: the probability the shuttle is on time, the average number of correct guesses on a quiz, the mean
wait for the next shuttle. Those fixed truths are **parameters** — properties of the model, not of any one
morning. This week we look at what happens when you actually *collect data* from such a model many times over
and summarize it. The summary you compute from data is a **statistic**, and the two limit theorems describe
how a particular statistic — the average — behaves as you collect more and more.

We anchor both demonstrations in one quantity from the recurring case: Maya's morning commute time. We model
it as

$$
C \sim \text{Normal}(\mu = 22,\ \sigma = 5),
$$

a Normal distribution with **mean** $\mu = 22$ minutes and **standard deviation** $\sigma = 5$ minutes (this
course always parameterizes the Normal by mean and standard deviation, never by variance — keep that in mind
when you call `rnorm`). The number $22$ is the parameter $E[C]$: the long-run average commute the model is
built around. It is *not* something you can read off a single morning; a single morning gives you one draw,
which might be $17$ or $29$ or anything else. The lab is about what many draws, properly summarized, reveal
about that hidden $22$.

Two questions drive the two simulations:

- **Law of large numbers (LLN).** If you average more and more commute times, does the average *settle*? The
  LLN says yes: the **sample mean** $\bar C_n = \frac{1}{n}\sum_{i=1}^{n} C_i$ converges to the **parameter**
  $E[C] = 22$ as $n$ grows. We will watch the running average flatten toward the dashed line at $22$.
- **Central limit theorem (CLT).** A single sample mean is itself a random number — collect a different week
  of mornings and you get a slightly different $\bar C_n$. The CLT describes how those sample means scatter:
  for large $n$, the distribution of $\bar C_n$ is approximately

$$
\bar C_n \ \approx\ \text{Normal}\!\left(\mu,\ \frac{\sigma}{\sqrt{n}}\right)
= \text{Normal}\!\left(22,\ \frac{5}{\sqrt{n}}\right).
$$

  We will draw thousands of sample means, histogram them, and check that the spread matches $5/\sqrt{n}$ and
  shrinks as $n$ grows.

The single most important habit this lab is training: **keep the statistic and the parameter distinct in your
head and in your words.** $\bar C_n$ is the sample mean, a thing you compute from data and that wiggles from
sample to sample. $E[C] = 22$ is the expectation, a fixed feature of the model that never wiggles. The LLN is
the statement that the first one homes in on the second; the CLT is the statement that the first one's wiggle
is Normal-shaped with a width you can predict. Everything below is in service of seeing that pair of facts.

![Left: the parameter $E[C]=22$ drawn as a single fixed point — it never moves. Right: ten independent
sample means, each computed from its own 30 simulated mornings, scattered around that same fixed value.
Synthetic; seed 35003.](../assets/figures/lab13_statistic_parameter_anchor.png){#fig-lab13-anchor
fig-alt="Two side-by-side number lines of commute time in minutes. The left panel shows one fixed black
dot at 22, labeled 'one fixed number, never wiggles.' The right panel shows ten scattered points between
about 20 and 23, straddling a dashed reference line at 22, labeled 'each mean wiggles.'"}

Look at the two panels before you run a single line of code. The left panel is the parameter — one number,
$22$, that does not change no matter how many times you re-draw the picture. The right panel is ten different
*statistics* — ten sample means, each built from its own batch of 30 simulated mornings — scattered above and
below the same dashed $22$. Every step below is really one long answer to the question those two panels raise:
how does the scatter on the right behave as you add more mornings per batch (Step 1), and as you collect more
batches (Steps 2–4)?

All data in this lab are **synthetic; seed 35003 is set** so that anyone running the code gets the same
stream. The code is shown for teaching and is not executed on this site (`#| eval: false`); run it yourself in
RStudio or Posit Cloud to reproduce the figures described.

## Goal

By the end of this lab you should be able to:

- Simulate a long stream of iid draws from $C \sim \text{Normal}(22, 5)$ with `rnorm` and a fixed seed.
- Compute a **running mean** with `cumsum(x) / seq_along(x)` and read its plot as a picture of the LLN.
- Generate many independent **sample means** of a fixed size $n$ and histogram them.
- Compare the **simulated** standard deviation of those means to the **theoretical** $5/\sqrt{n}$, and watch
  both fall as $n$ grows.
- Say, in one clean sentence each, what the LLN and the CLT are claiming — without confusing the sample mean
  (a statistic) with $E[C]$ (a parameter).

## Setup

You need only base R; no add-on packages are required. Work in a single Quarto file or R script so the whole
demonstration is reproducible end to end. Start every chunk that draws random numbers from the same seed so
your stream matches the one described here.

```{r}
#| label: lab13-setup
#| eval: false
# Base R only -- no packages needed.
set.seed(35003)

# The recurring-case model: Maya's commute time, in minutes.
# Normal parameterized by MEAN and STANDARD DEVIATION (the course convention).
mu    <- 22   # parameter E[C]: the long-run mean commute (a fixed truth of the model)
sigma <- 5    # parameter: the standard deviation of a single commute

# rnorm(n, mean, sd) takes the SD as its third argument -- not the variance.
# A quick sanity draw of five mornings:
rnorm(5, mean = mu, sd = sigma)
```

A note on the convention you just used: `rnorm`'s third argument is the **standard deviation**, which lines
up with how this course writes the Normal, $\text{Normal}(\mu, \sigma)$. If you ever see a formula written
with a variance $\sigma^2$, remember to pass $\sigma$ (here $5$), not $\sigma^2$ (which would be $25$). This
is exactly the kind of silent mix-up the notation ledger flags, so it is worth pausing on once.

Keep two names straight for the rest of the lab:

- `mu` (value $22$) stands for the **parameter** $E[C]$ — fixed, never estimated from the data below.
- the things you will compute with `mean(...)` stand for **statistics** — they are estimates that move from
  sample to sample. The whole point is to compare the second to the first.

## Steps

### Step 1 — Build one long stream and watch its running average settle (LLN)

The law of large numbers is about a *single* growing sample. Draw one long stream of commute times, then form
the running mean: the average of the first observation, then the first two, then the first three, and so on.
The compact base-R idiom is `cumsum(x) / seq_along(x)` — `cumsum(x)` gives the running totals and
`seq_along(x)` gives the counts $1, 2, 3, \dots$, so their ratio is the average so far at each step.

```{r}
#| label: lab13-lln
#| eval: false
set.seed(35003)
mu    <- 22
sigma <- 5

n_stream <- 5000                       # one long stream of mornings
x <- rnorm(n_stream, mean = mu, sd = sigma)

# Running mean: average of the first 1, first 2, first 3, ... observations.
running_mean <- cumsum(x) / seq_along(x)

# Look at how the running average evolves -- jumpy at the start, then settling.
head(running_mean, 10)                 # the first ten partial averages
tail(running_mean, 1)                  # the average over all 5000 mornings

# A picture of the LLN: the running mean vs. the number of mornings averaged.
plot(seq_along(running_mean), running_mean, type = "l",
     xlab = "number of mornings averaged (n)",
     ylab = "running sample mean of commute time (min)",
     main = "Law of large numbers: the average settles toward E[C] = 22")
abline(h = mu, lty = 2)                # the fixed parameter E[C] = 22
```

Read the plot left to right. On the far left, you have averaged only a handful of mornings, so a single
unusually long or short commute swings the average a lot — the curve is jagged and can sit well away from
$22$. As $n$ climbs into the hundreds and thousands, each new morning is one voice in a large chorus and moves
the average less and less; the curve flattens and hugs the dashed line at $\mu = 22$. That dashed line is the
**parameter** $E[C]$, the fixed target. The wiggling curve is the **statistic** $\bar C_n$, your sample mean
as a function of how much data you have folded in. The LLN is precisely the claim that the wiggling curve
homes in on the dashed line. Notice what it does *not* say: it does not say the average will ever equal $22$
exactly, and it does not say individual commute times stop varying — single mornings are as scattered as ever.
Only their *average* settles.

![The picture Step 1's code produces: the running mean of a single stream of 5,000 simulated commute times,
jagged for the first few hundred mornings and hugging $E[C]=22$ (dashed) from there on. Synthetic; seed
35003.](../assets/figures/lab13_lln_running_mean.png){#fig-lab13-lln
fig-alt="A line plot of running sample mean of commute time in minutes against number of mornings averaged,
from 1 to 5000. The line swings widely between about 21 and 28 for the first few hundred mornings, then
flattens and hugs a dashed horizontal reference line at 22 for the rest of the plot."}

At $n=10$ mornings the running mean sits at about $22.2$; by $n=1{,}000$ it is at about $21.9$; by the full
$n=5{,}000$ it has settled to about $21.9$ — each step closer to, and staying near, the dashed line. Because
the chunk fixes `set.seed(35003)`, running it in your own session reproduces this exact curve; what the LLN
guarantees in general (for any seed) is the *shape* — wild at first, then flat near $22$ — not this specific
sequence of numbers.

### Step 2 — Build a crowd of sample means of a fixed size (CLT, part one)

Now switch from one growing sample to *many* independent samples, each of the **same** fixed size $n$. From
each sample you keep only one number: its mean. Collecting many such means gives you a sample of sample means,
and the CLT is a statement about the shape of *that* collection.

Drawing the means efficiently is a small base-R trick: put $n \times m$ draws into a matrix with $n$ rows and
$m$ columns, so each column is one sample of size $n$, then take column means with `colMeans`.

```{r}
#| label: lab13-clt-means
#| eval: false
set.seed(35003)
mu    <- 22
sigma <- 5

n <- 30        # size of each sample (number of mornings per "week-and-a-half")
m <- 10000     # how many independent samples we draw

# Each COLUMN of this matrix is one independent sample of n commute times.
draws <- matrix(rnorm(n * m, mean = mu, sd = sigma), nrow = n, ncol = m)

# One sample mean per column -> a crowd of m sample means.
sample_means <- colMeans(draws)

length(sample_means)   # m = 10000 sample means
mean(sample_means)     # a statistic: the average OF the sample means (near 22)
head(sample_means, 5)  # five individual sample means -- each one wiggles around 22
```

Each entry of `sample_means` is one $\bar C_n$ — the average commute over $n = 30$ simulated mornings. No two
are identical, because each used a fresh draw of $30$ mornings; that is the sense in which the sample mean is
itself random. Their *average*, `mean(sample_means)`, sits very close to $22$, which is reassuring: the sample
mean is centered on the parameter it is estimating. But centering is only half the story — the CLT also tells
us the **shape** and the **spread** of this crowd, which is what the next step inspects.

![The first 60 of the 10,000 sample means this chunk produces, plotted as a jittered strip of points. Each
point is a whole batch of 30 mornings collapsed to one number. Synthetic; seed
35003.](../assets/figures/lab13_sample_means_strip.png){#fig-lab13-strip
fig-alt="A one-dimensional strip chart of 60 points scattered between about 20 and 24.5 along an axis
labeled sample mean of commute time over n = 30 mornings. The points cluster loosely around a dashed
reference line at 22, with a dotted line almost on top of it marking the average of all 10,000 sample
means, 22.003."}

This is the "crowd" the CLT is a statement about: not commute times themselves, but *sample means built from
30 commute times at a time*. The dashed line at $22$ and the dotted line at `mean(sample_means)` $\approx
22.003$ are nearly on top of each other — the crowd is centered right where the parameter says it should be.
What the picture does not yet show is how *tightly* the crowd clusters around that center; Step 3 puts a number
and a curve on exactly that.

### Step 3 — Histogram the means and compare the spread to the theory (CLT, part two)

Histogram the crowd of sample means and lay the theoretical curve over it. The CLT predicts that for a sample
of size $n$, the sample mean is approximately $\text{Normal}\!\left(22,\ 5/\sqrt{n}\right)$. The center should
be about $22$; the width — the standard deviation of the sample means, often called the **standard error** —
should be about $5/\sqrt{n}$. Compute the *simulated* spread directly with `sd(sample_means)` and compare it
to the *theoretical* $5/\sqrt{n}$.

```{r}
#| label: lab13-clt-hist
#| eval: false
set.seed(35003)
mu    <- 22
sigma <- 5
n <- 30
m <- 10000

draws        <- matrix(rnorm(n * m, mean = mu, sd = sigma), nrow = n, ncol = m)
sample_means <- colMeans(draws)

# The two numbers to compare:
theory_se <- sigma / sqrt(n)   # predicted spread of the sample means: 5 / sqrt(30)
sim_se    <- sd(sample_means)  # simulated spread, straight from the data

theory_se                      # what the CLT predicts
sim_se                         # what the simulation produced -- should be close

# Histogram of the sample means with the predicted Normal curve drawn on top.
hist(sample_means, breaks = 40, freq = FALSE,
     xlab = "sample mean of commute time over n = 30 mornings (min)",
     main = "CLT: sample means are ~ Normal(22, 5/sqrt(n))")
curve(dnorm(x, mean = mu, sd = theory_se), add = TRUE, lwd = 2)
abline(v = mu, lty = 2)        # the parameter E[C] = 22, where the crowd is centered
```

You should see a tidy bell sitting over $22$, with the overlaid Normal curve tracking the histogram closely,
and `sim_se` landing very near `theory_se`. The histogram is the empirical distribution of the **statistic**
$\bar C_n$; the smooth curve is the CLT's **predicted** distribution for it. When they match, you have *seen*
the central limit theorem rather than merely been told it.

![The histogram this chunk produces (10,000 sample means, $n=30$ each) with the predicted
$\text{Normal}(22,\ 5/\sqrt{30})$ curve overlaid. Synthetic; seed
35003.](../assets/figures/lab13_clt_hist_overlay.png){#fig-lab13-clt-hist
fig-alt="A histogram of sample means of commute time over 30 mornings, bell-shaped and centered near 22,
with a smooth Normal density curve overlaid that closely tracks the bar heights, and a dashed vertical
line at 22."}

Numerically, `theory_se` $=5/\sqrt{30}\approx 0.913$ and `sim_se` $\approx 0.917$ in this run — close enough
that the smooth curve and the bars in the figure are almost indistinguishable. That agreement, not just the
bell shape, is the CLT's real content: it predicts the *width* of the crowd, not only that the crowd is
bell-shaped.

(In this particular lab the per-morning commute is already Normal, so the sample means are essentially exactly
Normal at every $n$ — a clean case for reading off the spread. The deeper claim of the CLT, which you can
explore on your own, is that the bell shape emerges *even when the thing you average is not Normal*, as long as
$n$ is reasonably large. Try replacing `rnorm` with draws from a skewed shape and watch the crowd of means
turn bell-shaped anyway.)

![A transfer check on that claim, using a generic skewed distribution (Exponential with rate 1, mean 1 —
illustrative only, not the commute model), not `rnorm`. Left: a single draw is heavily skewed, nothing like a
bell. Right: the mean of 30 such draws is already close to bell-shaped, tracking a
$\text{Normal}(1,\ 1/\sqrt{30})$ curve. Synthetic; seed
35003.](../assets/figures/lab13_skewed_clt_practice.png){#fig-lab13-skewed
fig-alt="Two histograms side by side. The left one, of single draws from a skewed exponential distribution,
is tall near zero and decays to the right with a long right tail. The right one, of means of 30 such draws,
is roughly bell-shaped and centered near 1, with a dashed Normal curve tracking its bars closely."}

Notice what stayed the same and what changed: the *shape you average* (badly skewed, on the left) has nothing
to do with the *shape of the average* (bell-shaped, on the right) once $n=30$ is large enough. That is the
CLT's universality — it does not require the per-observation distribution to be Normal, only that you are
averaging enough of them. (This particular skewed example is a generic illustration for this exercise, not
part of the recurring commuter's-morning model.)

### Step 4 — Shrink the spread by growing the sample size

The CLT's $5/\sqrt{n}$ has a $\sqrt{n}$ in the denominator, so larger samples give *narrower* crowds of means:
more mornings per sample means each sample mean is a better-pinned estimate of $22$. Loop over a few sample
sizes and watch the simulated spread fall in lockstep with the theoretical $5/\sqrt{n}$.

```{r}
#| label: lab13-shrink
#| eval: false
set.seed(35003)
mu    <- 22
sigma <- 5
m     <- 10000

for (n in c(5, 30, 100, 400)) {
  draws        <- matrix(rnorm(n * m, mean = mu, sd = sigma), nrow = n, ncol = m)
  sample_means <- colMeans(draws)

  theory_se <- sigma / sqrt(n)
  sim_se    <- sd(sample_means)

  cat("n =", n,
      "| theoretical 5/sqrt(n) =", round(theory_se, 3),
      "| simulated sd of means =", round(sim_se, 3),
      "| center =", round(mean(sample_means), 3), "\n")
}
```

As $n$ goes from $5$ to $400$, the theoretical spread $5/\sqrt{n}$ falls from $5/\sqrt{5} \approx 2.24$ down to
$5/\sqrt{400} = 0.25$, and the simulated `sd` of the means tracks it the whole way. The center stays glued to
$22$ throughout — growing $n$ does not move the target, it only sharpens the aim. That single denominator is
why "more data" buys you precision: to halve the spread of your average you need *four times* as many
observations, because $\sqrt{n}$, not $n$, is doing the work.

![The printed table above, drawn as a picture: theoretical $5/\sqrt{n}$ (solid) against the simulated
standard deviation of the sample means (dashed) at $n=5,30,100,400$. Synthetic; seed
35003.](../assets/figures/lab13_se_shrink_plot.png){#fig-lab13-shrink
fig-alt="A line plot of standard error of the sample mean in minutes against sample size n, for n = 5, 30,
100, and 400. Two nearly overlapping curves, one theoretical and one simulated, both fall steeply from
about 2.2 at n=5 to about 0.25 at n=400."}

The two curves sit almost exactly on top of each other at every $n$ — the simulated spread is not just falling,
it is falling at the rate the formula predicts. Notice the curve is steep between $n=5$ and $n=30$ and much
flatter between $n=100$ and $n=400$: the *first* handful of extra mornings buys you far more precision than the
same number of extra mornings does later on, because it is $\sqrt{n}$, not $n$, in the denominator.

## Verify

Run the four steps in your own session (with `set.seed(35003)`) and check each of these. They are quick,
ungraded sanity checks — if any one fails, re-read the step it belongs to before moving on.

- **The stream is centered right.** In Step 1, `tail(running_mean, 1)` (the average over all $5000$ mornings)
  should sit very close to $22$. If it is far off, confirm you passed `sd = 5` to `rnorm` and not `sd = 25`.
- **The running mean settles, it does not freeze.** The Step 1 plot should be jagged on the left and flat on
  the right, hugging the dashed line at $22$ — not a flat line from the start (that would mean you forgot the
  running average) and not still wild on the right (that would mean too short a stream).

  ![Both named failure modes, rendered so you can recognize them if your own plot looks like this. Left: the
  bug of forgetting the running average entirely and instead plotting the single overall mean, flat from the
  very first point. Right: the bug of stopping the stream too early ($n=60$ instead of $5{,}000$), so the curve
  is still visibly wild at the right edge. Synthetic; seed
  35003.](../assets/figures/lab13_running_mean_diagnostic_panel.png){#fig-lab13-diagnostic
  fig-alt="Two line plots side by side, both of a quantity called running average against number of
  mornings averaged. The left panel shows a jagged solid line settling toward a dashed reference at 22, with
  a second, dashed line lying perfectly flat at that same level from the very start — the flat-from-the-start
  bug. The right panel shows a single short, still-swinging line that stops at n = 60, never reaching the flat
  part seen in the correct plot."}

  If your own Step 1 plot looks like the flat dashed line in the left panel, you likely computed the overall
  `mean(x)` once and repeated it, instead of `cumsum(x) / seq_along(x)`. If it looks like the short, still-wild
  line in the right panel, your stream is too short to show settling — increase `n_stream`.
- **The crowd is centered on the parameter.** In Step 2, `mean(sample_means)` should be near $22$. This is the
  statistic (the average of your sample means) closing in on the parameter $E[C] = 22$.
- **Simulated spread matches theory.** In Step 3, `sim_se` and `theory_se` should agree to about two decimal
  places. For $n = 30$ both should be near $5/\sqrt{30} \approx 0.913$.
- **The spread shrinks like $5/\sqrt{n}$.** In Step 4, the two columns of numbers should move together as $n$
  grows, and the center should stay near $22$ for every $n$.
- **You can say it cleanly.** Finish by writing one sentence for each theorem. A good LLN sentence: *"The
  sample mean $\bar C_n$ converges to the parameter $E[C] = 22$ as $n$ grows."* A good CLT sentence: *"For
  large $n$, the sample mean is approximately $\text{Normal}(22,\ 5/\sqrt{n})$, narrowing as $n$ grows."*
  Notice that both sentences name $\bar C_n$ as the moving statistic and $22$ as the fixed parameter — that
  distinction is the whole lesson.

## AI use note

If you use an AI assistant while working through this lab, treat it as a study partner whose output you always
check against the running code and the printed numbers — never as the source of truth. Disclose use where your
course or assignment asks you to. Everything an assistant suggests here is verifiable in seconds by running the
chunk and reading the result, so verify before you trust.

| Tool | Purpose | Verification |
|---|---|---|
| AI coding assistant | Explain what `cumsum(x) / seq_along(x)` computes step by step | Run it on a tiny vector like `1:4` and confirm the partial averages by hand |
| AI coding assistant | Suggest a base-R way to histogram the sample means and overlay a Normal curve | Run the chunk; check the curve tracks the bars and the histogram centers on 22 |
| AI chat assistant | Restate the LLN and CLT in plain words | Compare its statement to the Week 13 note; confirm it keeps the sample mean (statistic) distinct from $E[C]$ (parameter) and uses spread $5/\sqrt{n}$ |
| AI chat assistant | Explain why the spread of the means is $5/\sqrt{n}$ and not $5/n$ | Recompute `sd(sample_means)` at two sample sizes and check it falls like $1/\sqrt{n}$, not $1/n$ |

## See also

- [Week 13 — Sums, simulation & limit behavior](../notes/week-13-sums-simulation-limit-behavior.qmd) — the
  companion note stating the LLN and CLT in words and symbols.
- [Lab 02 — Monte Carlo basics](lab-02-monte-carlo-basics.qmd) — where the simulate-many-times-and-summarize
  habit started; this lab is its limit-theorem payoff.
- [Distribution reference](../resources/distribution-reference.qmd) — the Normal's mean–standard-deviation
  parameterization and how it maps to `rnorm`.
- [R and Quarto setup](../resources/r-quarto-setup.qmd) — getting a reproducible session running with
  `set.seed` so your stream matches this page.

*The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.*
