---
title: "Lab 9 — Simulating discrete models"
subtitle: "Generating draws from binomial and Poisson models and comparing to the formulas"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab puts the named discrete models from
> [Week 9 — Common discrete models](../notes/week-09-common-discrete-models.qmd) on the computer. There you
> learned to read a probability, a mean, and a variance straight off a model's parameters. Here you *generate*
> draws from those same models and check, by counting, that the empirical frequencies land where the formulas
> say they should. Seeing the two agree — formula and simulation — is the point.

## The idea

In the Week 9 note you committed to a habit: name the model, state its assumptions, then read its
probabilities off the parameters. A binomial count of correct guesses, a Poisson count of shuttle arrivals —
each came with an exact pmf, an exact mean, and an exact variance. That note did the arithmetic by hand.

A simulation does something complementary. Instead of summing pmf terms, it *manufactures* the random
variable: it asks R to play the experiment out tens of thousands of times and then counts how often each
result occurred. If your model is right, the long-run **relative frequency** of an event settles toward the
probability the formula gives, and the **average** and **spread** of the simulated draws settle toward the
model's mean and variance. This is the same Monte Carlo logic you met in
[Lab 2 — Monte Carlo basics](lab-02-monte-carlo-basics.qmd), now aimed at two named models rather than a
single hand-built experiment.

Why bother, when you already have the exact answers? Three reasons. First, agreement between an independent
simulation and a hand calculation is a genuine check on both — a typo in the formula or a misread parameter
usually shows up as a mismatch. Second, the simulation makes the *variability* tangible: you watch four
arrivals an hour fail to be exactly four most of the time, instead of just being told the variance is four.
Third, when a problem gets too tangled to sum by hand, simulation is often the only practical route to an
answer — so it is worth trusting it on cases you *can* check by hand first. That is exactly what this lab
does: every simulated quantity has a formula answer sitting next to it, so you can see the method working
before you ever rely on it blind.

A note on what the simulation can and cannot tell you. A simulated relative frequency is itself a random
quantity — run it again with a different seed and you get a slightly different number. It *approximates* the
true probability, and the approximation gets tighter as you draw more samples, but it never lands exactly on
the formula value. So "close" is the standard you are checking against, not "identical." Part of the skill
this lab builds is developing a feel for how close is close enough, and for why a hundred thousand draws gives
a steadier answer than a hundred.

## Goal

By the end of this lab you should be able to:

- Generate many independent draws from a binomial model with `rbinom()` and from a Poisson model with
  `rpois()`, with a fixed seed so the run is reproducible.
- Estimate a tail probability, a mean, and a variance from simulated draws by counting and averaging, and
  compare each estimate to the exact formula value from the Week 9 note.
- Confirm the binomial quiz result $P(X \ge 8) \approx 0.0547$, the mean of $5$, and the variance of $2.5$ by
  simulation, and the Poisson result $P(N = 4) \approx 0.195$ with mean and variance both near $4$.
- Explain why R's `*geom` family uses a **different geometric convention** than the course (failures before
  the first success, support starting at $0$) and adjust for it.
- Judge whether a simulated estimate is "close enough" to its formula value, and say what makes it closer.

All data in this lab are **synthetic; seed 35003 is set** in every simulation chunk so your run reproduces.

## Setup

You need a working R session with Quarto — see
[R · Quarto setup](../resources/r-quarto-setup.qmd) if you have not set one up yet. Everything here is base R;
no packages to install. The four functions you will lean on are part of R's built-in distribution family:

- `rbinom(n, size, prob)` — draw `n` binomial counts, each the number of successes in `size` independent
  trials with success probability `prob`.
- `rpois(n, lambda)` — draw `n` Poisson counts, each with rate (mean) `lambda`.
- `mean()` and `var()` — the sample average and sample variance of a vector of draws.
- a logical comparison such as `x >= 8` or `y == 4`, which R turns into `TRUE`/`FALSE`; wrapping it in
  `mean()` gives the *proportion* that are `TRUE`, i.e. an estimated probability.

That last trick is the engine of the whole lab, so it is worth pausing on. In R, `TRUE` counts as $1$ and
`FALSE` as $0$. So `mean(x >= 8)` adds up one for every draw that clears $8$ and divides by how many draws
there were — exactly the relative frequency of the event $\{X \ge 8\}$. Estimating a probability becomes a
one-line `mean()` of a comparison.

Two quantities to fix before you start, both straight from the Week 9 note. The quiz model is
$X \sim \text{Binomial}(10, 0.5)$ — ten true/false questions, each guessed correctly with probability
$0.5$. The arrivals model is $N \sim \text{Poisson}(4)$ — shuttle arrivals in an hour at a rate of four per
hour. Keep the formula targets in view as you go, because the whole exercise is comparison:

$$
P(X \ge 8) = \frac{56}{1024} \approx 0.0547, \qquad E[X] = 5, \qquad \operatorname{Var}(X) = 2.5,
$$

$$
P(N = 4) = \frac{e^{-4}\,4^{4}}{4!} \approx 0.195, \qquad E[N] = \operatorname{Var}(N) = 4.
$$

Pick a sample size large enough to make the relative frequencies steady. This lab uses $100{,}000$ draws
throughout; smaller runs work but wobble more, which is itself something worth seeing (the last step asks you
to try it).

## Steps

The R chunks below are **shown for study** — they carry `#| eval: false`, so they are not executed on this
site. Read each chunk, make sure you can say what every line does, then run it yourself in your own R session.
Every chunk sets `set.seed(35003)` so your numbers reproduce.

Every step below walks the same four-box path: start from the formula, simulate a hundred thousand draws,
summarize them, and compare. Keeping that shape in mind makes each step a variation on one habit rather than
four unrelated recipes.

![The four-step path every step in this lab follows: formula, simulate, summarize, compare. Synthetic; seed
35003.](../assets/figures/lab09_sim_formula_pipeline.png){#fig-lab09-pipeline fig-alt="A left-to-right flow
diagram of four boxes connected by arrows: Formula (pmf, mean, variance); Simulate (rbinom or rpois, 100,000
draws); Summarize (mean, var, mean of a logical comparison); Compare (close?)."}

> This diagram is not a new step — it names the shape that Steps 1 through 4 already share. Wherever a step
> asks you to compare a simulated number to a formula value, it is walking these same four boxes.

### Step 1 — Draw from the binomial quiz model and check mean and variance

Start with the quiz. Generate one hundred thousand independent plays of the ten-question quiz, where each play
returns the number of correct guesses out of ten. Then check that the simulated average and variance match the
formula values $5$ and $2.5$.

```{r}
#| label: binom-draw-moments
#| eval: false
set.seed(35003)

n_sim <- 100000

# each draw is the number correct in 10 fair-guess true/false questions
x <- rbinom(n_sim, size = 10, prob = 0.5)

# the simulated distribution of scores, as a quick sanity check
table(x)

# compare to the formula values from the Week 9 note
mean(x)   # should sit close to E[X]  = n*p       = 5
var(x)    # should sit close to Var(X)= n*p*(1-p) = 2.5
```

A few things to notice in the output. `table(x)` shows the *shape* of the binomial — counts bunched around
$5$, thinning out toward $0$ and $10$, symmetric because $p = 0.5$. The `mean(x)` should print something very
close to $5$ and `var(x)` something very close to $2.5$. They will not be exactly $5$ and $2.5$: these are
sample statistics from a finite run, so expect agreement to two or three figures, not perfect equality. If
`mean(x)` came back near $2.5$ instead of $5$, that would be a red flag — almost certainly a wrong `size` or
`prob`, not bad luck. The numbers being *in the right neighborhood* is the check; the small wobble is normal.

![The 100,000 simulated quiz scores (bars) against the exact pmf (line and dots), with the run's own
`mean(x)` and `var(x)` shown against the formula values $5$ and $2.5$. Synthetic; seed
35003.](../assets/figures/lab09_step1_binom_hist.png){#fig-lab09-step1-hist fig-alt="Bar chart of the relative
frequency of 100,000 simulated ten-question quiz scores, symmetric and peaked at 5, with the exact binomial
pmf plotted as a line of dots lying almost exactly on top of the bars; mean 5.008 and variance 2.493 are
printed above the peak against the formula values 5 and 2.5."}

> **What this shows.** The bars (what the simulation produced) and the line (what the formula predicts) sit
> almost exactly on top of each other at every score from $0$ to $10$ — that overlap *is* the check `table(x)`,
> `mean(x)`, and `var(x)` are asking you to make. The non-visual equivalent is the run's own printed numbers:
> `mean(x)` and `var(x)` next to the formula's $5$ and $2.5$.

### Step 2 — Estimate the binomial tail probability $P(X \ge 8)$

Now reproduce the headline result from the note: the chance a pure guesser scores at least eight out of ten.
The formula gives $56/1024 \approx 0.0547$. The simulation estimates it by counting the fraction of draws that
clear eight.

```{r}
#| label: binom-tail
#| eval: false
set.seed(35003)

n_sim <- 100000
x <- rbinom(n_sim, size = 10, prob = 0.5)

# P(X >= 8): the fraction of plays scoring 8, 9, or 10
mean(x >= 8)          # should sit close to 56/1024 = 0.0547

# break it down to see where the mass sits
mean(x == 8)          # ~ 45/1024 ~ 0.0439
mean(x == 9)          # ~ 10/1024 ~ 0.0098
mean(x == 10)         # ~  1/1024 ~ 0.0010

# and confirm the exact target the estimate is chasing
56 / 1024             # 0.0546875
```

The line `mean(x >= 8)` is the whole estimate: it asks, of one hundred thousand plays, what fraction scored
eight or higher, and that fraction should land near $0.0547$. Breaking it into `x == 8`, `x == 9`, and
`x == 10` shows the same structure the formula used — the $45$, $10$, and $1$ arrangements over $1024$ — now
appearing as three relative frequencies that add up to the tail. Because $P(X \ge 8)$ is a small probability,
it is the noisiest estimate in this lab: a rare event needs more draws to pin down than a common one, which is
why a hundred thousand plays (not a hundred) is worth the wait here. Confirm your simulated value rounds to
about $0.055$ and you have matched the note by an independent route.

![The exact pmf (bars, with $x \ge 8$ highlighted) against the simulated relative frequency (open circles) at
every score. Synthetic; seed 35003.](../assets/figures/lab09_step2_tail_compare.png){#fig-lab09-step2-tail
fig-alt="The same binomial bar chart with the three rightmost bars at 8, 9, and 10 highlighted and braced,
labeled with the exact tail sum 56/1024 approximately 0.0547; open circles for the simulated relative
frequency sit on every bar including the highlighted ones, and a label reports the simulated estimate
approximately 0.0546."}

> **What this shows.** The highlighted bars are the *same* three bars ($8$, $9$, $10$) the note summed by
> hand; the open circles are where $100{,}000$ simulated plays actually landed. Formula ($\approx 0.0547$) and
> simulation ($\approx 0.0546$) land within a thousandth of each other — the kind of agreement Step 2 asks you
> to confirm without a picture, here made visible.

### Step 3 — Draw from the Poisson arrivals model and check $P(N = 4)$ and the mean–variance equality

Switch to the shuttle-arrivals model, $N \sim \text{Poisson}(4)$. Generate one hundred thousand hours of
arrivals, estimate the chance of exactly four arrivals in an hour, and confirm the Poisson signature that mean
and variance are *both* near the rate $\lambda = 4$.

```{r}
#| label: pois-draw
#| eval: false
set.seed(35003)

n_sim <- 100000

# each draw is the number of shuttle arrivals in one hour at rate lambda = 4
y <- rpois(n_sim, lambda = 4)

# the simulated distribution of hourly counts
table(y)

# P(N = 4): the fraction of hours with exactly four arrivals
mean(y == 4)          # should sit close to e^-4 * 4^4 / 4! ~ 0.195

# the Poisson signature: mean and variance both near lambda = 4
mean(y)               # ~ 4
var(y)                # ~ 4

# the exact target the P(N=4) estimate is chasing
exp(-4) * 4^4 / factorial(4)   # 0.1953668
```

The `mean(y == 4)` line estimates $P(N = 4)$ the same way Step 2 estimated the binomial tail — the fraction of
draws equal to four — and it should land near $0.195$. Notice that even though four per hour is the *average*,
exactly four happens under a fifth of the time; the rest of the probability spreads over three, five, two,
six, and beyond. That spread is the variance at work. The pair `mean(y)` and `var(y)` should *both* sit near
$4$: this near-equality is the Poisson's fingerprint, and seeing it in simulated data is a quick way to gauge
whether a real count dataset is plausibly Poisson. If the simulated variance came out much larger than the
mean — say near $10$ while the mean stayed near $4$ — that would signal the steady-independent-rate story was
violated, exactly the diagnostic the note described.

![The 100,000 simulated hourly arrival counts (bars, with $y = 4$ in iron) against the exact Poisson(4) pmf
(dashed line and dots). Synthetic; seed 35003.](../assets/figures/lab09_step3_poisson_hist.png){#fig-lab09-step3-hist
fig-alt="Bar chart of the relative frequency of 100,000 simulated hourly shuttle-arrival counts, right-leaning
and peaked near 3 to 4, with the bar at exactly 4 colored iron and labeled with the simulated estimate
approximately 0.1950 against the exact value approximately 0.1954; the exact Poisson(4) pmf is overlaid as a
dashed line of dots; mean 4.009 and variance 3.992 are printed above, both close to lambda = 4."}

> **What this shows.** The highlighted bar at $y = 4$ is the single tallest bar, but most of the shaded area
> belongs to the *other* bars — a picture of "four is the average, not the guarantee." The overlaid dashed
> curve is the exact pmf; the fact that it sits almost on top of the simulated bars is the same agreement
> `mean(y == 4)` reports as a number, and `mean(y)` next to `var(y)` is the Poisson signature (both near $4$)
> read straight off the plot's title numbers.

### Step 4 — Mind the geometric convention, and feel the role of sample size

This step has two short parts, both about *not* being misled.

**Part A — R's geometric counts differently than the course.** The Week 9 note fixed the course's geometric as
the **number of trials up to and including the first success**, so its support starts at $1$. R's `*geom`
family — `rgeom`, `dgeom`, `pgeom` — instead counts the **number of failures *before* the first success**, so
its support starts at $0$. The two differ by exactly one. If you simulate a "waiting for the first success"
count with `rgeom` and forget this, every value is one too small and your estimated mean comes out as
$\frac{1}{p} - 1$ instead of $\frac{1}{p}$. The fix is to add one:

```{r}
#| label: geom-convention
#| eval: false
set.seed(35003)

n_sim <- 100000
p <- 0.25

# R's rgeom: FAILURES before the first success, support starts at 0
r_failures <- rgeom(n_sim, prob = p)
mean(r_failures)          # ~ (1 - p)/p = 3   (this is NOT the course's mean)

# convert to the COURSE convention: trials up to and INCLUDING the first success
trials <- r_failures + 1  # now support starts at 1
mean(trials)              # ~ 1/p = 4         (this matches the note's E[X] = 1/p)
```

The takeaway is not arithmetic; it is a habit. Before you trust any simulated quantity, confirm the function's
convention matches the course's. `rbinom` and `rpois` happen to agree with the note's parameterizations, so
Steps 1–3 needed no adjustment; `rgeom` does not, so it needs the `+ 1`. The
[Distribution reference](../resources/distribution-reference.qmd) lists each model's R function and flags this
geometric shift explicitly.

![R's `rgeom()` convention (failures before the first success, support starting at $0$, left panel) against
the course convention (trials up to and including the first success, support starting at $1$, right panel),
both at $p = 0.25$. Synthetic; seed 35003.](../assets/figures/lab09_step4a_geom_shift.png){#fig-lab09-step4a-geom
fig-alt="Two side-by-side bar charts of the same decaying geometric shape. Left panel: R's rgeom convention,
support starting at 0, with a dashed mean line at 3 labeled mean of r_failures approximately 3.02 against
(1-p)/p = 3. Right panel: the course convention, the identical shape shifted right by one, support starting
at 1, with a dashed mean line at 4 labeled mean of trials approximately 4.02 against 1/p = 4."}

> **What this shows.** The two panels are the *identical shape*, shifted by exactly one trial — that shift is
> the whole convention pitfall in one picture. The left panel's mean lands near $3$ (R's own convention); only
> after adding $1$ does the right panel's mean land near $4$, matching the course's $E[X] = 1/p$. If you ever
> see a simulated geometric mean sitting one unit low, this is the picture to check against.

**Part B — sample size controls the wobble.** Re-run the binomial tail estimate from Step 2 at a few different
sample sizes and watch the estimate steady up as the sample grows. Smaller runs scatter more widely around
$0.0547$; larger runs hug it.

```{r}
#| label: sample-size-wobble
#| eval: false
set.seed(35003)

target <- 56 / 1024   # 0.0546875, the exact value the estimates chase

for (n_sim in c(100, 1000, 10000, 100000)) {
  x <- rbinom(n_sim, size = 10, prob = 0.5)
  est <- mean(x >= 8)
  cat("n =", n_sim, " estimate =", round(est, 4),
      " error =", round(est - target, 4), "\n")
}
```

At $n = 100$ the estimate can be off in the second decimal place — sometimes there are simply not enough
high-scoring plays to pin a rare event down. By $n = 100{,}000$ it should track $0.0547$ closely. This is the
law-of-large-numbers idea you will meet head-on in
[Lab 13 — Law of large numbers and CLT](lab-13-law-of-large-numbers-and-clt.qmd): more draws, steadier
estimate. For now it explains *why* this lab used a hundred thousand draws — small simulations are honest but
noisy, and a tail probability needs the larger run to be convincing.

![Two hundred replicate estimates of $P(X \ge 8)$ at each sample size (boxes), the page's own single
seed-35003 run overlaid (diamonds), and the exact target $56/1024$ (dashed line). Synthetic; seed
35003.](../assets/figures/lab09_step4b_wobble.png){#fig-lab09-step4b-wobble fig-alt="Four boxplots of the
estimate mean of x greater than or equal to 8 at sample sizes 100, 1,000, 10,000, and 100,000, with the box
width and whisker spread shrinking sharply from left to right toward a dashed horizontal line at the exact
target approximately 0.0547; a diamond marker on each box shows where the page's own single seed-35003 run
landed at that sample size."}

> **What this shows.** Each box is two hundred *independent* replays of the exact same experiment at one
> sample size — its height is the wobble the page's text describes. At $n = 100$ the box spans roughly $0.01$
> to $0.12$, wide enough to occasionally miss the target badly (the diamond there, from the page's own
> seed-35003 run, is a low outlier); by $n = 100{,}000$ the box has collapsed to a sliver hugging the dashed
> target line. The diamonds are exactly the four numbers the shown `sample-size-wobble` chunk would print if
> you ran it — the boxes are the added picture of how much *another* seed could have differed.

## Verify

Line up each simulated quantity against the formula value from the Week 9 note and confirm they agree to a
sensible number of figures. You are checking for *closeness*, not exact equality — a finite simulation always
carries a little noise.

| Quantity | Formula value (Week 9) | What the simulation should show |
|---|---|---|
| $E[X]$, binomial quiz | $np = 5$ | `mean(x)` near $5$ |
| $\operatorname{Var}(X)$, binomial quiz | $np(1-p) = 2.5$ | `var(x)` near $2.5$ |
| $P(X \ge 8)$ | $56/1024 \approx 0.0547$ | `mean(x >= 8)` near $0.055$ |
| $P(N = 4)$, Poisson arrivals | $e^{-4}4^{4}/4! \approx 0.195$ | `mean(y == 4)` near $0.195$ |
| $E[N]$ and $\operatorname{Var}(N)$ | both $= 4$ | `mean(y)` and `var(y)` both near $4$ |
| Geometric mean, course convention | $1/p = 4$ at $p = 0.25$ | `mean(rgeom(...) + 1)` near $4$ |

If every row matches to two or three figures, the simulation and the formulas have confirmed each other, and
you can trust the method on a problem where you *cannot* sum by hand. If a row is badly off — not just noisy
but in the wrong neighborhood — treat it as a signal, not a fluke: re-check the `size`, `prob`, or `lambda`
you passed, confirm you used the right comparison (`>=` vs `==`), and for any geometric work confirm you added
the `+ 1`. A good final habit: re-run a chunk with a *different* seed and confirm the estimate moves only a
little. A quantity that swings wildly between seeds at $n = 100{,}000$ is telling you something is wrong with
the setup, not with chance.

## AI use note

If you use an AI assistant while working through this lab, keep a short record like the one below. The
load-bearing column is **Verification** — how *you* confirmed the output was right, independent of the tool.
An AI can produce R that runs cleanly and still computes the wrong thing; the formula targets in this lab are
exactly what let you catch that.

| Tool | Purpose | Verification |
|---|---|---|
| AI chat assistant | Explain why `mean(x >= 8)` estimates a probability | Re-derived it by hand: `TRUE` counts as $1$, so the mean is the fraction of draws clearing $8$; confirmed it tracked $56/1024 \approx 0.0547$ |
| AI chat assistant | Draft the `rpois` chunk for the arrivals model | Checked `mean(y)` and `var(y)` both landed near $4$ and `mean(y == 4)` near $0.195$ before trusting the code; confirmed `lambda = 4` was passed |
| AI coding helper | Remind me of R's geometric convention | Did not take the answer on faith — ran `mean(rgeom(n, p))` and saw it gave $(1-p)/p$, confirming the `+ 1` shift to the course's mean $1/p$ |

If you used no AI for a step, write "none" — an honest empty record is the correct record.

## See also

- [Week 9 — Common discrete models](../notes/week-09-common-discrete-models.qmd) — the companion note, where
  the binomial and Poisson pmfs, means, variances, and the $P(X \ge 8)$ and $P(N = 4)$ results are derived by
  hand.
- [Lab 2 — Monte Carlo basics](lab-02-monte-carlo-basics.qmd) — the earlier lab that introduced estimating a
  probability by repeated trials and counting, the engine this lab reuses.
- [Distribution reference](../resources/distribution-reference.qmd) — each model's pmf, mean, variance,
  support, and R function in one table, including the geometric support shift between R and the course.
- [R · Quarto setup](../resources/r-quarto-setup.qmd) — how to get a reproducible R + Quarto session running
  before you start.

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
