---
title: "Lab 13 — Law of large numbers & CLT"
subtitle: "Watching averages stabilize and sums go normal"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** Run, by hand and on screen, the two limit results stated in the companion
> [Week 13 note](../notes/week-13-sums-simulation-limit-behavior.qmd): the **law of large numbers** (the
> average of many commute times settles onto the true mean) and the **central limit theorem** (the
> distribution of that average is approximately Normal). You will generate synthetic data with a fixed
> seed, plot what happens as the sample size grows, and check the simulated summaries against the exact
> values the theory predicts. The code is base R, shown for study, and reproducible — you run it yourself.

## The idea

All term you have *derived* probabilities by hand: counting outcomes, integrating densities, applying
Bayes' rule. This lab adds the simulation half of the course. Instead of proving the week's two theorems,
you generate data that obey them and **watch the results appear**. That habit — define a synthetic model,
fix a seed, simulate many trials, and compare the simulated summary to a theoretical prediction — is the
same engine you will reuse in the Week-14 project, so treat the code here as a reusable skeleton rather
than a one-off.

Two questions sit underneath the whole lab. *If I average many noisy measurements, does the average settle
down to the truth?* And *if it does, what shape does the leftover wobble take?* The first is answered by the
**law of large numbers** (LLN): for independent and identically distributed (iid) draws $X_1, X_2, \dots$
with mean $\mu$, the sample mean converges to $\mu$,

$$
\bar X_n \;=\; \frac{1}{n}\sum_{i=1}^{n} X_i \;\longrightarrow\; \mu \qquad \text{as } n \to \infty.
$$

The second is answered by the **central limit theorem** (CLT): the sample mean of $n$ iid draws is
approximately Normal, centered at $\mu$ with a standard deviation that shrinks like $\sigma/\sqrt n$,

$$
\bar X_n \;\approx\; \operatorname{Normal}\!\left(\mu, \; \frac{\sigma}{\sqrt n}\right)
\qquad \text{for large } n.
$$

The LLN explains why averages are **reliable** — they home in on the right number. The CLT explains why
they are **approximately Normal** — the wobble around that number has a predictable bell shape whose width
is set by $\sigma/\sqrt n$. We are using this lab's recurring world because it makes both claims checkable
against an *exact* target.

## Goal

Use simulation to make the LLN and CLT visible for Maya's morning commute, and confirm the simulated
numbers match the exact theory from the [Week 13 note](../notes/week-13-sums-simulation-limit-behavior.qmd).
Concretely, by the end you will have:

1. Plotted one **running average** of commute times and watched it settle onto the true mean of $22$
   minutes — the LLN.
2. Built the **sampling distribution** of the average of $n$ commutes, histogrammed it, and overlaid the
   Normal curve the CLT predicts — confirming the center is $22$ and the spread is $5/\sqrt n$.
3. **Varied $n$** to see the bell tighten by the $\sqrt n$ factor exactly as the formula says.

The recurring case makes the comparison clean: because the commute model is *already* Normal, the CLT
prediction is not just an approximation here but the exact distribution of the average, so any mismatch
between simulation and theory points to a bug, not to "the approximation hasn't kicked in yet."

## Setup

*All data here are **synthetic; seed set** (`set.seed(35003)`). The R chunks are **shown for study**,
written in base R, and **not executed** on this site (`#| eval: false`) — you run them in your own R
session. See the [R · Quarto setup](../resources/r-quarto-setup.qmd) page if you need to get RStudio or
Posit Cloud going first.*

The world is the one from Week 11: Maya's morning commute time is

$$
C \sim \operatorname{Normal}(\mu = 22 \text{ min}, \; \sigma = 5 \text{ min}),
$$

where $\mu = 22$ is the **mean** and $\sigma = 5$ is the **standard deviation** (not a variance — this is
the single most common numeric trap of the week). Each simulated morning is one independent draw from this
model. You only need three quantities defined up front, and the course seed so your run matches the note's:

```{r}
#| label: lab13-setup
#| eval: false
# Lab 13 setup. Synthetic commute model C ~ Normal(mean = 22, sd = 5). Seed is the course seed.
set.seed(35003)
mu    <- 22     # true mean commute time (minutes)
sigma <- 5      # true standard deviation (minutes) -- a SD, NOT a variance
# A quick sanity draw: ten commutes from the model. They should scatter around 22 by a few minutes.
rnorm(10, mean = mu, sd = sigma)
```

Read every `sd =` argument in this lab as "standard deviation." If you ever pass the variance ($\sigma^2 =
25$) by mistake, nothing errors — your "commutes" just scatter five times too wide and every later check
fails quietly. With `mu`, `sigma`, and the seed in hand, work the steps below in order; each one builds on
the last.

## Steps

### Step 1 — Watch one running average settle (the LLN)

Draw a long stream of commute times and plot the *cumulative* average after $1, 2, 3, \dots$ mornings. The
cumulative average after $n$ mornings is $\bar C_n = (C_1 + \cdots + C_n)/n$, which `cumsum()` divided by
the running count computes in one line. Early on, with only a handful of mornings, the average lurches
around — one unusually long commute moves it a lot. As $n$ grows, each new morning is a smaller fraction of
the total, the average steadies, and it hugs the line $\mu = 22$.

```{r}
#| label: lab13-lln-running-average
#| eval: false
#| fig-cap: "Running average of simulated commute times converging to the true mean of 22 minutes."
#| fig-alt: "A jagged line that smooths and settles onto a dashed horizontal line at 22 as the number of mornings increases."
# Watch the running mean of commute times settle to mu = 22 as n grows (the law of large numbers).
set.seed(35003)
mu    <- 22
sigma <- 5
n     <- 5000
C        <- rnorm(n, mean = mu, sd = sigma)   # n synthetic commute times
run_mean <- cumsum(C) / seq_len(n)            # cumulative average after 1, 2, ..., n mornings

plot(seq_len(n), run_mean, type = "l",
     xlab = "number of mornings (n)", ylab = "running average commute (min)",
     main = "Law of large numbers: running mean -> 22")
abline(h = mu, lty = 2)                        # the true mean the average is converging to

# The average at a few sample sizes -- it should drift toward 22 and stay there.
run_mean[c(10, 100, 1000, 5000)]
```

What you are meant to see: a ragged line for small $n$ that flattens onto the dashed $\mu = 22$ line as $n$
grows. That flattening *is* the LLN. Notice the line never freezes perfectly on $22$ — it keeps a tiny
wobble — but the wobble shrinks, and at $n = 5000$ the running average should sit within a few hundredths of
$22$. Before moving on, look at the printed `run_mean` values: $n = 10$ may be off by a minute or so, while
$n = 5000$ should be very close to $22$.

### Step 2 — Build the sampling distribution of the mean (the CLT)

The LLN tracked *one* average growing. The CLT is about the *distribution of the average across many
repeats of the experiment*. So now repeat the whole experiment many times: take $n = 30$ commutes, average
them to get one value $\bar C_{30}$, and do that for thousands of independent weeks-worth of mornings. The
workhorse is `replicate()`, which runs an expression many times and collects the results into a vector —
here a vector of thousands of independent sample means. Histogram them and overlay the Normal the CLT
predicts.

```{r}
#| label: lab13-clt-sampling-distribution
#| eval: false
#| fig-cap: "Histogram of many sample means overlaid with the matching Normal(22, 5/sqrt(30)) curve."
#| fig-alt: "A bell-shaped histogram of sample means centered at 22, closely matched by an overlaid normal curve."
# Form many independent sample means of n = 30 commutes each, then histogram them (the CLT).
set.seed(35003)
mu    <- 22
sigma <- 5
reps  <- 10000                                  # number of independent sample means to form
n     <- 30                                      # commutes averaged per sample mean
means <- replicate(reps, mean(rnorm(n, mean = mu, sd = sigma)))

hist(means, breaks = 40, freq = FALSE,
     xlab = "sample mean of 30 commutes (min)",
     main = "CLT: distribution of the sample mean")
# Overlay the Normal the CLT predicts: center mu = 22, sd = sigma / sqrt(n) = 5 / sqrt(30).
curve(dnorm(x, mean = mu, sd = sigma / sqrt(n)), add = TRUE, lwd = 2)

# Summaries to compare against the theory in the Verify section.
c(observed_mean = mean(means),                   # should be ~ 22
  observed_sd   = sd(means),                      # should be ~ 5 / sqrt(30) ~ 0.913
  predicted_sd  = sigma / sqrt(n))
```

What you are meant to see: a tidy bell sitting right on top of the overlaid Normal curve, centered near
$22$ with a spread near $5/\sqrt{30}$. The overlaid curve uses $\operatorname{sd} = \sigma/\sqrt n$, **not**
$\sigma$ — feeding it the raw $\sigma = 5$ is a frequent slip that produces a curve five-and-a-half times
too wide for the histogram. Because the commute model is already Normal, the histogram should match the
curve almost exactly; that tight match is your signal the code is right.

### Step 3 — Vary n and watch the bell tighten by √n

The formula says the width of the sampling distribution is $\sigma/\sqrt n$, so quadrupling $n$ should
*halve* the spread, and multiplying $n$ by nine should cut it to a third. Test that directly: rebuild the
sampling distribution for several values of $n$ and put the histograms side by side on a common axis so the
tightening is visible at a glance.

```{r}
#| label: lab13-clt-vary-n
#| eval: false
#| fig-cap: "Histograms of sample means for n = 5, 30, 120 on a common axis, each matched by its Normal(22, 5/sqrt(n)) curve."
#| fig-alt: "Three bell-shaped histograms that get visibly narrower as the sample size n increases, each overlaid with a matching normal curve."
# Rebuild the sampling distribution of the mean for several n and watch the bell tighten like sigma/sqrt(n).
set.seed(35003)
mu    <- 22
sigma <- 5
reps  <- 10000
ns    <- c(5, 30, 120)                            # quadrupling n should roughly halve the spread

par(mfrow = c(1, 3))                              # three panels side by side
for (n in ns) {
  means <- replicate(reps, mean(rnorm(n, mean = mu, sd = sigma)))
  hist(means, breaks = 40, freq = FALSE,
       xlim = c(16, 28),                           # common axis so the panels are comparable
       xlab = "sample mean (min)",
       main = paste("n =", n))
  curve(dnorm(x, mean = mu, sd = sigma / sqrt(n)), add = TRUE, lwd = 2)
  # Report observed vs predicted spread for this n.
  cat("n =", n,
      " observed_sd =", round(sd(means), 3),
      " predicted_sd =", round(sigma / sqrt(n), 3), "\n")
}
par(mfrow = c(1, 1))
```

What you are meant to see: three bells, all centered on $22$ (the LLN guarantees the center never moves),
but progressively *narrower* as $n$ grows. The center staying put while the spread shrinks is the LLN and
the CLT in one picture — the average is reliable (it lands on $22$) and its wobble is a predictable,
tightening bell. The printed `observed_sd` values should track the `predicted_sd` values closely; the next
section turns that "closely" into a numeric check.

## Verify

A simulation is only convincing if it lands on a number you can compute independently. Here you can,
because the commute model is Normal, so the CLT prediction is **exact**: the average of $n$ iid commutes is
distributed *exactly* $\bar C_n \sim \operatorname{Normal}(22, 5/\sqrt n)$ for every $n$. Use these three
checks.

**1. The running average lands on the mean (Step 1).** The printed `run_mean[5000]` should sit within a few
hundredths of the exact value

$$
\mu \;=\; 22 \text{ minutes.}
$$

If $n = 5000$ is still off by a full minute, suspect a bug — most likely a missing or misplaced
`set.seed()`, or a `sd` argument that was handed the variance.

**2. The sampling-distribution spread matches $\sigma/\sqrt n$ (Step 2).** With $n = 30$ the exact
predicted standard deviation of the average is

$$
\operatorname{sd}(\bar C_{30}) \;=\; \frac{\sigma}{\sqrt n} \;=\; \frac{5}{\sqrt{30}}
\;\approx\; \frac{5}{5.477} \;\approx\; 0.913 \text{ minutes,}
$$

and the exact predicted center is $22$. The reported `observed_mean` should be near $22$ and the
`observed_sd` near $0.913$. A handy gut check: averaging just $30$ mornings cuts the day-to-day spread from
$5$ minutes down to about $0.91$ — a roughly five-fold tightening, which is the $\sqrt{30} \approx 5.48$
factor the formula promises.

**3. The √n scaling holds across $n$ (Step 3).** For the three sizes the exact predicted spreads are

$$
\frac{5}{\sqrt 5} \approx 2.236, \qquad
\frac{5}{\sqrt{30}} \approx 0.913, \qquad
\frac{5}{\sqrt{120}} \approx 0.456 \text{ minutes.}
$$

Notice the $\sqrt n$ scaling directly: going from $n = 30$ to $n = 120$ (a four-fold increase in sample
size) halves the predicted spread from $\approx 0.913$ to $\approx 0.456$. Each printed `observed_sd`
should match its `predicted_sd` to about two decimals. If all three
checks agree, the simulation and the theory are telling the same story, which is the entire point of running
the lab. If they disagree, re-read the two snags in the Week-13 note's debugging section — a missing seed,
or a variance passed where R wanted a standard deviation, accounts for nearly every mismatch.

## AI use note

If you use an AI assistant on this lab, record it briefly and honestly. The **Verification** column is the
load-bearing one: an AI can produce R that runs cleanly and still computes the wrong quantity (overlaying a
Normal with the wrong standard deviation is a classic), so a hand check against the theory above is
non-negotiable. *(This table shows the format and is not an endorsement of any tool — disclose your actual
use.)*

| Tool | Purpose | Verification |
|---|---|---|
| AI chat assistant | Explain why `replicate()` returns a vector of sample means rather than one number | Re-read `?replicate`; confirmed `length(means)` equals `reps` and that each entry is one `mean(rnorm(n, ...))` |
| AI chat assistant | Suggest base-R syntax to overlay a Normal curve on a histogram | Checked the overlaid `dnorm` used $\sigma/\sqrt n$, not $\sigma$; confirmed the reported `observed_sd` matched the exact $5/\sqrt{30}\approx 0.913$ |
| AI code helper | Draft the cumulative-average line for the LLN plot in Step 1 | Verified `cumsum(C)/seq_len(n)` equals a manual average at $n = 10$ by hand before trusting the plot |

The standing rule: any number an AI helps produce must be checked against a hand calculation or the exact
theory in the Verify section before you rely on it. Disclose AI use on graded work as Blackboard directs.

## See also

- [Week 13 — Sums, simulation & limit behavior](../notes/week-13-sums-simulation-limit-behavior.qmd) — the
  companion note that states and motivates the LLN and CLT.
- [Week 14 — Probability modeling project](../notes/week-14-probability-modeling-project.qmd) — where this
  define-model / set-seed / simulate / verify pattern becomes the project workflow.
- [R · Quarto setup](../resources/r-quarto-setup.qmd) — getting RStudio or Posit Cloud running.
- [Notation glossary](../resources/notation-glossary.qmd)
- [Distribution reference](../resources/distribution-reference.qmd)

The graded deliverable, its rubric, and due date live in Blackboard (the LMS) — this page is study and
practice only.
