---
title: "Week 13 — Sums, simulation & limit behavior"
subtitle: "Seeing the law of large numbers and the central limit theorem by simulation"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
readings:
  - "Grinstead & Snell, Ch 7 — Sums of Random Variables — supports the variance-of-a-sum rule and the setup"
  - "Grinstead & Snell, Ch 8 — Law of Large Numbers — supports the LLN statement"
  - "Grinstead & Snell, Ch 9 — Central Limit Theorem — supports the CLT statement"
  - "MIT OCW 18.05 — central limit theorem and the bootstrap/simulation strand — supports the simulation picture"
toc: true
---

This is a **software week**. The whole term we have *derived* probabilities by hand — counting outcomes,
integrating densities, applying Bayes' rule. This week we add the other half of the course's promise:
*simulation*. We will not prove the week's two big results; we will state them and then **watch them
happen** by generating synthetic data in R and counting. The page is built on the
`software_note_lab` anatomy — a concept note, a practiced simulation sequence, a reproducibility
convention, a debugging note, and an honest record of any AI help — and it hands off to the companion
[Lab 13](../labs/lab-13-law-of-large-numbers-and-clt.qmd) where you run the code yourself.

## Concept note

Two questions sit underneath almost everything you do with data. *If I average many noisy measurements,
does the average settle down to the truth?* And *if it does, what does the leftover wobble look like?*
This week names the two theorems that answer them — the **law of large numbers** (LLN) and the
**central limit theorem** (CLT) — and uses simulation to make them visible. Before either, we need one
small piece of algebra about adding random variables.

### Adding random variables, and the variance of a sum

Expectation adds with no fine print. For any random variables $X$ and $Y$,

$$
E[X + Y] \;=\; E[X] + E[Y],
$$

whether or not $X$ and $Y$ are independent. Variance is the careful one. The general rule carries a
cross-term:

$$
\operatorname{Var}(X + Y) \;=\; \operatorname{Var}(X) + \operatorname{Var}(Y) + 2\,\operatorname{Cov}(X, Y).
$$

The covariance $\operatorname{Cov}(X, Y)$ is exactly the joint-dependence quantity from Week 12 — it
measures whether $X$ and $Y$ move together. When $X$ and $Y$ are **independent** (or merely uncorrelated)
the covariance is $0$, the cross-term vanishes, and variance becomes additive:

$$
\operatorname{Var}(X + Y) \;=\; \operatorname{Var}(X) + \operatorname{Var}(Y)
\qquad (X \perp Y).
$$

That special case is the engine of the whole week. Stack up $n$ **independent and identically
distributed** (iid) copies $X_1, X_2, \dots, X_n$, each with mean $\mu$ and standard deviation $\sigma$,
and look at their sum and their average. For the sum $S_n = X_1 + \cdots + X_n$,

$$
E[S_n] \;=\; n\mu,
\qquad
\operatorname{Var}(S_n) \;=\; n\sigma^2,
$$

and for the sample mean $\bar X_n = S_n / n$,

$$
E[\bar X_n] \;=\; \mu,
\qquad
\operatorname{Var}(\bar X_n) \;=\; \frac{\sigma^2}{n},
\qquad
\operatorname{sd}(\bar X_n) \;=\; \frac{\sigma}{\sqrt n}.
$$

Read that last line slowly, because it is the seed of both theorems. The average is **centered on the
truth** $\mu$ for every $n$, and its spread **shrinks** like $\sigma/\sqrt n$ as $n$ grows. The mean stays
put; the wobble around it gets smaller.

### The law of large numbers — the average settles

The LLN turns "the spread shrinks" into a promise about the average itself. As the sample size grows, the
sample mean of iid draws converges to the true expectation:

$$
\bar X_n \;=\; \frac{1}{n}\sum_{i=1}^{n} X_i \;\longrightarrow\; \mu \;=\; E[X]
\qquad \text{as } n \to \infty.
$$

In words: pile up enough independent observations and their average gets and stays close to the value you
would have computed if you knew the distribution exactly. This is the formal license behind a habit you
already have — estimating a probability or a mean by *running many trials and averaging*. Every Monte
Carlo estimate in this course (the lab-2 counting, the lab-5 posterior) is the LLN cashing out: the
long-run proportion converges to the true probability because a proportion is just the average of $0/1$
indicators. We state the LLN; we do not prove it. Its plausibility comes straight from the previous
section — if $\operatorname{Var}(\bar X_n) = \sigma^2/n \to 0$, the average has less and less room to stray
from $\mu$.

### The central limit theorem — the wobble is a bell

The LLN says the average lands on $\mu$. The CLT describes the *shape* of how it scatters along the way,
and the answer is remarkable: almost regardless of the shape of the original distribution, the
**standardized** sum of many iid draws looks Normal. Precisely,

$$
Z_n \;=\; \frac{S_n - n\mu}{\sigma\sqrt n}
\;=\; \frac{\bar X_n - \mu}{\sigma / \sqrt n}
\;\longrightarrow\; \operatorname{Normal}(0, 1)
\qquad \text{as } n \to \infty.
$$

Undo the standardizing and the same statement says the sample mean of $n$ iid draws is **approximately
Normal**, centered at $\mu$ with standard deviation $\sigma/\sqrt n$:

$$
\bar X_n \;\approx\; \operatorname{Normal}\!\left(\mu, \; \frac{\sigma}{\sqrt n}\right)
\qquad \text{for large } n.
$$

This is why the Normal distribution is everywhere: anything built by *adding up many small independent
contributions* — measurement error, total wait across many stops, an average of many quiz scores — tends
toward a bell curve, even when each contribution is not itself Normal. We state the CLT; we do not prove
it. The point of the week is that you can *see* both results on screen in a few lines of R, without taking
a limit by hand.

A convenient feature of this week's recurring case: when the draws are *already* Normal, the
approximation is exact. If $C \sim \operatorname{Normal}(22, 5)$, then for **every** $n$ the average of $n$
iid commutes is exactly $\bar C_n \sim \operatorname{Normal}(22, 5/\sqrt n)$. That makes the commute the
perfect first thing to simulate — we know the exact target to compare against — and then we will switch to
a decidedly non-Normal source to show the CLT bending it into a bell anyway.

## Setup and practice sequence

*All data below are **synthetic; seed set** (`set.seed(35003)`). The R chunks are **shown for study**,
written in base R, and **not executed** on this site (`#| eval: false`) — you run them yourself in the
[companion lab](../labs/lab-13-law-of-large-numbers-and-clt.qmd).* Work the sequence in order; each step
builds on the last.

The recurring world: Maya's morning commute time is $C \sim \operatorname{Normal}(\mu = 22, \sigma = 5)$
minutes, the same model from Week 11. We will (1) watch one running average settle to $22$ — the LLN; (2)
build the sampling distribution of $\bar C_n$ and see it is the bell the CLT promises; and (3) repeat the
CLT step on a *non-Normal* source — a sum of quiz scores — to show the bell appears even when the
ingredients are not Normal.

**1. Watch one running average settle (the LLN).** Draw a long stream of commute times and plot the
*cumulative* average after $1, 2, 3, \dots$ mornings. Early on the average jumps around; as $n$ grows it
hugs the line $\mu = 22$.

```{r}
#| label: wk13-lln-running-average
#| eval: false
#| fig-cap: "Running average of simulated commute times converging to the true mean of 22 minutes."
#| fig-alt: "A jagged line that smooths and settles onto a dashed horizontal line at 22 as the number of mornings increases."
# Synthetic commute times C ~ Normal(mu = 22, sd = 5). Watch the running mean settle to 22 (LLN).
set.seed(35003)
n  <- 5000
C  <- rnorm(n, mean = 22, sd = 5)        # NOTE: sd = 5 is a standard deviation, NOT a variance
run_mean <- cumsum(C) / seq_len(n)       # cumulative average after 1, 2, ..., n mornings

plot(seq_len(n), run_mean, type = "l",
     xlab = "number of mornings (n)", ylab = "running average commute (min)",
     main = "Law of large numbers: running mean -> 22")
abline(h = 22, lty = 2)                  # the true mean mu the average is converging to

run_mean[c(10, 100, 1000, 5000)]         # the average at a few sample sizes: drifting toward 22
```

What you are meant to see: the line is ragged for small $n$ and flattens onto the dashed $\mu = 22$ line as
$n$ grows. That flattening *is* the LLN, and the rate it tightens is the $\sigma/\sqrt n$ shrinkage from the
concept note.

**2. Build the sampling distribution of the mean (the CLT).** Now repeat the *experiment* many times.
Take $n = 30$ commutes, average them to get one $\bar C_{30}$, and do that for many independent weeks-worth
of mornings. Histogram the resulting averages.

```{r}
#| label: wk13-clt-sampling-distribution
#| eval: false
#| fig-cap: "Histogram of many sample means of 30 commute times, overlaid with the Normal(22, 5/sqrt(30)) curve."
#| fig-alt: "A bell-shaped histogram of sample means centered at 22, closely matched by an overlaid normal curve."
# Many independent sample means of n = 30 commutes each. The histogram should look Normal (CLT),
# centered at 22 with sd = 5/sqrt(30).
set.seed(35003)
reps <- 10000                            # number of independent sample means to form
n    <- 30                               # commutes averaged per sample mean
means <- replicate(reps, mean(rnorm(n, mean = 22, sd = 5)))

hist(means, breaks = 40, freq = FALSE,
     xlab = "sample mean of 30 commutes (min)",
     main = "CLT: distribution of the sample mean")
# overlay the Normal the CLT predicts: center 22, sd 5/sqrt(30)
curve(dnorm(x, mean = 22, sd = 5 / sqrt(30)), add = TRUE, lwd = 2)

c(observed_mean = mean(means),           # should be ~ 22
  observed_sd   = sd(means),             # should be ~ 5/sqrt(30) ~ 0.913
  predicted_sd  = 5 / sqrt(30))
```

What you are meant to see: a tidy bell sitting on top of the overlaid Normal curve, its center near $22$
and its spread near $5/\sqrt{30} \approx 0.913$. Because the source was already Normal, the match is
essentially exact here — a clean baseline before we stress-test the CLT next.

**3. Stress-test the CLT on a non-Normal sum (a sum of quiz scores).** Here is the surprising part. Take a
clearly non-Normal ingredient — a single true/false quiz score, which is *binomial*, not a bell — and look
at the **sum across several quizzes**. The CLT says the total will drift toward Normal as we add more.

```{r}
#| label: wk13-clt-nonnormal-sum
#| eval: false
#| fig-cap: "Histograms of sums of quiz scores for k = 1, 4, 20, approaching a normal shape as more quizzes are added."
#| fig-alt: "Three histograms of quiz-score totals; the leftmost is lumpy and the rightmost is a smooth bell, each matched by a normal curve."
# Each quiz score X ~ Binomial(10, 0.5) (mean 5, sd ~ 1.58), which is NOT Normal on its own.
# Sum scores across k quizzes and watch the total's distribution turn into a bell as k grows (CLT).
set.seed(35003)
reps <- 10000
draw_total <- function(k) sum(rbinom(k, size = 10, prob = 0.5))   # total over k quizzes

par(mfrow = c(1, 3))                      # compare k = 1, 4, 20 side by side
for (k in c(1, 4, 20)) {
  totals <- replicate(reps, draw_total(k))
  hist(totals, breaks = 30, freq = FALSE,
       xlab = paste("sum of", k, "quiz score(s)"),
       main = paste("k =", k))
  curve(dnorm(x, mean = 5 * k, sd = sqrt(2.5 * k)), add = TRUE, lwd = 2)  # CLT target
}
par(mfrow = c(1, 1))
```

What you are meant to see: at $k = 1$ the histogram is the chunky, slightly blocky binomial; by $k = 4$ it
is already bell-ish; by $k = 20$ it is a clean Normal hugging the overlaid curve with mean $5k$ and
standard deviation $\sqrt{2.5k}$ (using the quiz's $\mu = 5$, $\sigma^2 = 2.5$ from Week 8). The ingredient
never changed — the **sum** is what goes Normal. That is the CLT in one picture.

### Two worked reads of the output

*Worked read — the recurring commuter's slice.* In step 2 the prediction is fixed in advance:
$\bar C_{30} \approx \operatorname{Normal}(22,\, 5/\sqrt{30})$, so

$$
\operatorname{sd}(\bar C_{30}) \;=\; \frac{5}{\sqrt{30}} \;\approx\; \frac{5}{5.477} \;\approx\; 0.913
\text{ minutes.}
$$

So averaging just $30$ mornings cuts the day-to-day spread from $\sigma = 5$ minutes down to about $0.91$
minutes — a roughly five-fold tightening, exactly the $\sqrt{30} \approx 5.48$ factor the formula predicts.
The simulation's reported `observed_sd` should land right on that $0.913$; if it does, the code and the
theory agree, which is the whole point of running it.

*Transfer read — sums of quiz scores.* In step 3, one quiz has $\mu = 5$ and $\sigma^2 = 2.5$ (the
Week-8 binomial numbers). Because the $k$ quizzes are independent, expectation and variance both add:

$$
E\!\left[\sum_{i=1}^{k} X_i\right] = 5k,
\qquad
\operatorname{Var}\!\left(\sum_{i=1}^{k} X_i\right) = 2.5k,
\qquad
\operatorname{sd}\!\left(\sum_{i=1}^{k} X_i\right) = \sqrt{2.5k}.
$$

For $k = 20$ that is a center of $5(20) = 100$ and a standard deviation of $\sqrt{2.5 \cdot 20} =
\sqrt{50} \approx 7.07$ — the very mean and sd fed to the overlaid `dnorm` curve. The histogram matching
that curve is the CLT making a Normal out of a stack of non-Normal pieces.

## Reproducible-file convention

A simulation result that nobody else can reproduce is an anecdote, not evidence. The convention for every
piece of code on this page and in the lab is small and strict.

- **One `.qmd` per analysis.** Keep the narrative, the code, and the output in a single Quarto document so
  the reasoning and the computation never drift apart. You knit the same file to share a result.
- **Seed at the top, always.** Put `set.seed(35003)` *before* the first random draw in any chunk that uses
  randomness. The seed makes the pseudo-random stream deterministic: re-knit the file and you get the
  identical histogram and the identical reported numbers. Without it, every run differs and "I got
  $0.913$" becomes unverifiable. (We use the same course seed, $35003$, everywhere so results line up
  across the term.)
- **End with `sessionInfo()`.** Close the document with a chunk that records your R version and loaded
  package versions, so a reader can tell *which* environment produced the numbers.

```{r}
#| label: wk13-session-info
#| eval: false
# Record the environment at the end of every reproducible analysis file.
sessionInfo()
```

Together — one file, a fixed seed, and a recorded session — these three habits mean another person (or
future-you) can re-run the document and get exactly your figures. That is what "reproducible" means in
this course, and it is the standard the Week-14 project will be held to.

## Debugging

Two snags account for most "my simulation looks wrong" moments this week. Both are quiet — the code runs
without an error, it just gives the wrong thing — so they are worth recognizing on sight.

- **Forgetting `set.seed()` → non-reproducible output.** If you omit the seed (or place it *after* the
  random draw, which is the same as omitting it), every knit produces a different histogram and different
  reported numbers. The code is not broken, but the result cannot be checked against anyone else's. *Fix:*
  put `set.seed(35003)` as the first line that runs before any `rnorm`/`rbinom`/`replicate`. If you want to
  confirm a result is *stable* rather than a seed artifact, deliberately try a couple of different seeds and
  confirm the picture and the summary numbers barely move — for a large simulation they should not.
- **`rnorm` takes a standard deviation, not a variance.** This is the single most common numeric bug of the
  week, and it ties straight back to the course's notation. In R, `rnorm(n, mean = 22, sd = 5)` expects the
  third argument to be the **standard deviation** $\sigma = 5$. If you instead pass the *variance*
  $\sigma^2 = 25$, writing `rnorm(n, mean = 22, sd = 25)`, you have silently quintupled the spread — your
  "commute times" now scatter by $25$ minutes, your running average settles much more slowly, and your CLT
  histogram is five times too wide. Nothing errors; the numbers are just wrong. *Fix:* read every `sd =`
  argument as "standard deviation," and sanity-check against the prediction — if step 2 reports an
  `observed_sd` near $4.6$ instead of $0.913$, you have fed it a variance. (The same trap lurks in `dnorm`
  and `pnorm`, which also take `sd`, never a variance.)

## AI Use Note

If you use an AI assistant on the simulation work, record it briefly and honestly — the **Verification**
column is the one that matters, because an AI can produce R that runs cleanly and still computes the wrong
quantity. *(This table is an example of the format, not an endorsement of any tool; disclose actual use.)*

| Tool | Purpose | Verification |
|---|---|---|
| AI chat assistant | Explain why `replicate()` returns a vector of sample means rather than one number | Re-read R help `?replicate`; confirmed the length equals `reps` by checking `length(means)` |
| AI chat assistant | Suggest base-R syntax to overlay a Normal curve on a histogram | Checked the overlaid `dnorm` used $\mu$ and $\sigma/\sqrt n$, not $\sigma$; confirmed `observed_sd` matched $5/\sqrt{30}\approx 0.913$ |
| AI code helper | Draft the cumulative-average line for the LLN plot | Verified `cumsum(C)/seq_len(n)` equals a manual average at $n=10$ by hand before trusting the plot |

The standing rule: any number an AI helps produce must be checked against a hand calculation or the
theory from the notes before you rely on it. Disclose AI use on graded work as Blackboard directs.

## Reading and source pointer

This week is grounded in **Grinstead & Snell** (GNU FDL, free online): **Chapter 7 — Sums of Random
Variables** for adding random variables and the variance-of-a-sum rule, **Chapter 8 — Law of Large
Numbers** for the LLN statement, and **Chapter 9 — Central Limit Theorem** for the CLT statement. The
simulation-as-evidence emphasis — watching a sampling distribution form on screen — echoes the
computational strand in **MIT OCW 18.05** (CC BY-NC-SA 4.0). *These notes are the course's own synthesis,
grounded in but not copied from the sources.* The commute model, the quiz-sum example, the R code, and all
numbers are original to this course; data are synthetic with the seed set to $35003$. We **state** the LLN
and CLT here and do not prove them; the proofs live in the source chapters for the curious.

## Public vs. graded

These notes, the examples, and the practice here are **public and ungraded** — study material only. No
graded prompts, answer keys, rubrics, point values, or due dates appear on this site. Graded
checkpoints, quizzes, homework, labs, the midterm, the project, and the final live in **Blackboard
(the LMS)**, which is authoritative for due dates, submissions, and grades. If this page and Blackboard
ever disagree, follow Blackboard.

## Portfolio connection

The pattern you practiced this week — *define a synthetic model, fix a seed, simulate many trials, and
compare the simulated summary to a theoretical prediction* — is exactly the engine of the **Week-14
probability-modeling project**. There the modeling question is "how often does Maya have $\ge 2$ late days
in a $5$-day week?", and the workflow is the same three moves you just ran: state the model, simulate it
reproducibly, and check the Monte Carlo estimate against a by-hand probability. The LLN is your license to
trust the estimate when the number of trials is large; the reproducible-file convention (one `.qmd`, a set
seed, a recorded session) is the standard the project's deliverable will be expected to meet. Treat this
week's three chunks as the reusable skeleton you will adapt in Week 14.

## See also

- [Lab 13 — Law of large numbers and CLT](../labs/lab-13-law-of-large-numbers-and-clt.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)
- [Distribution reference](../resources/distribution-reference.qmd)
- [Course syllabus](../syllabus.qmd)
