---
title: "Week 11 — Common continuous models"
subtitle: "Uniform, exponential, and normal"
page_mode: normal_week
week_contract: public_note
source_mode: source-informed
readings:
  - "Grinstead & Snell, Ch 5 — Important Distributions (continuous) — supports the catalogue of named continuous models and when each applies"
  - "MIT OCW 18.05 — review of continuous distributions and the normal — supports the standardization and 68–95–99.7 framing"
toc: true
---

## The week question

Last week you learned the *grammar* of continuous random variables: a density $f(x)$ whose area gives
probability, a cdf $F(x) = P(X \le x)$, and the rule that any single point has probability zero. That
grammar is general — it works for *any* continuous variable. This week we stock the shelf with the three
named densities that cover most of what you will actually meet, just as Week 9 stocked the discrete shelf.
The week's question is again a **recognition** question: **given a continuous scenario, which standard
model fits, and what feature of the story tells you so?** Once you can name the model — Uniform,
Exponential, or Normal — its density, its mean, and the way you read probabilities off it all come along
for free.

We have already met the machinery these models plug into. In Week 10 you computed probabilities as areas
($P(a \le X \le b) = \int_a^b f(x)\,dx$) and built cdfs by integrating densities. This week we do not
re-derive that machinery; we name the three densities you will reach for most, say what each one assumes,
and practice matching a story to a model.

## Why this matters

Naming the model is what turns a sentence like "the shuttle comes on average every fifteen minutes" into a
number you can compute. If you can say "the wait is $\text{Exponential}$ with rate $4$ per hour," you
immediately know its density, that its mean is $15$ minutes, and how to get $P(\text{wait} \le 15)$ — no
fresh integration setup required each time. As in Week 9, the hard part is rarely the arithmetic; it is
reading the assumptions in the story carefully enough to choose the right model.

These three models also travel far beyond shuttles. The same **exponential** that describes a waiting time
describes the lifetime of a component before it fails, the gap between earthquakes, or the time until a
radioactive atom decays. The same **normal** that describes a commute time describes measurement error,
heights in a population, or the average of many small independent effects (a fact Week 13 will make precise
with the Central Limit Theorem). And the **uniform** — "any value in a range, all equally likely" — is the
honest model for genuine no-information situations and the engine behind nearly every simulation you will
run. Learning the *assumptions* behind each, not just its formula, is what lets you reuse them.

## Learning goals

By the end of this week you should be able to:

- State, in words and in symbols, the Uniform, Exponential, and Normal models, including each density and
  (as results cross-linked to Weeks 8 and 10) each mean.
- Read a continuous scenario and identify which of the three standard models its assumptions select — and
  name the feature that rules the others out.
- Compute a probability for our recurring commuter case: an exponential shuttle wait and a normal commute
  time, using the cdf and the standard-normal $\Phi$.
- Apply the **68–95–99.7 rule** and the **standardization** $Z = (X - \mu)/\sigma$ to read normal
  probabilities.
- Recognize a *new* scenario — the time until a component fails — as exponential, and explain the
  **memoryless** property that makes it special.
- Read shown R code (`dunif`/`punif`, `dexp`/`pexp`, `dnorm`/`pnorm`) so you can see how software returns
  these same probabilities.

## Core vocabulary

- **Uniform$(a, b)$** — a continuous variable equally likely to land anywhere in the interval $[a, b]$; its
  density is flat. Written $X \sim \text{Uniform}(a, b)$.
- **Exponential$(\lambda)$** — the waiting time until the next event when events arrive at a constant
  **rate** $\lambda$ (the Poisson rate from Week 9). Written $T \sim \text{Exponential}(\lambda)$; its mean
  is $1/\lambda$.
- **Rate $\lambda$** — for the exponential, the average number of events per unit time; large $\lambda$
  means short waits. Mind the parameterization: we use **rate**, not mean, so $\text{Exponential}(4)$ has
  mean $1/4$.
- **Normal$(\mu, \sigma)$** — the symmetric bell-shaped model parameterized by its **mean** $\mu$ and its
  **standard deviation** $\sigma$. Written $X \sim \text{Normal}(\mu, \sigma)$.
- **Standard normal** $Z$ — the special case $\text{Normal}(0, 1)$; its cdf is written $\Phi(z) = P(Z \le z)$.
- **Standardization** — the transform $Z = (X - \mu)/\sigma$ that re-expresses any normal value as a number
  of standard deviations from the mean, so a single $\Phi$ table or function handles every normal.
- **Memoryless property** — the exponential's signature: having already waited does not change the
  distribution of the remaining wait.

## Concept development

### Uniform: every value in a range equally likely

The simplest continuous model is the **Uniform$(a, b)$**. The variable can land anywhere in $[a, b]$, and
no part of that interval is favored over another, so the density must be **flat**. Because the total area
under a density is $1$ and the interval has width $b - a$, the height is forced to be $1/(b - a)$:

$$
f(x) = \frac{1}{b - a}, \quad a \le x \le b, \qquad f(x) = 0 \text{ otherwise.}
$$

Probabilities are then just proportions of the interval — a rectangle's area. For $a \le c \le d \le b$,

$$
P(c \le X \le d) = \frac{d - c}{b - a},
$$

and by symmetry the mean sits in the middle, $E[X] = (a + b)/2$. Use the uniform when the honest
description is "somewhere in this range, with no reason to prefer any part of it": a rounding error in
$[-0.5, 0.5]$, a random arrival time within a fixed window, or — crucially — the output of a computer's
random-number generator, which is the raw material every simulation in this course is built from.

### Exponential: waiting time at a steady rate

In Week 9 we counted *how many* shuttles arrive in an hour with a Poisson at rate $\lambda = 4$. Now ask
the companion question: *how long until the next one?* When events arrive at a constant rate $\lambda$ and
do not coordinate, the waiting time $T$ is **Exponential$(\lambda)$**, with density

$$
f(t) = \lambda e^{-\lambda t}, \quad t \ge 0, \qquad f(t) = 0 \text{ for } t < 0.
$$

Integrating this density (the Week 10 move) gives a particularly clean cdf, which is what you actually use
to get probabilities:

$$
F(t) = P(T \le t) = 1 - e^{-\lambda t}, \quad t \ge 0.
$$

Its mean is the reciprocal of the rate,

$$
E[T] = \frac{1}{\lambda},
$$

which is exactly the intuition "rate $4$ per hour means about one every $1/4$ hour." The exponential's
defining feature is the **memoryless property**: if you have already waited $s$ minutes, the probability
of waiting at least $t$ more is the same as the original probability of waiting at least $t$ — the variable
has no memory of time already spent. Symbolically,

$$
P(T > s + t \mid T > s) = P(T > t).
$$

This is what makes the exponential the natural model for "pure waiting at a steady rate": a shuttle that is
already five minutes late is, under this model, no more "due" than a fresh wait. (Whether that is *true* of
real shuttles is exactly the kind of assumption you should question — but it is what the model says.)

### Normal: the bell curve and standardization

The **Normal$(\mu, \sigma)$** is the symmetric, bell-shaped model that shows up whenever a quantity is the
sum of many small independent influences — measurement noise, biological variation, the average of many
trials. Its density is

$$
f(x) = \frac{1}{\sigma \sqrt{2\pi}}\, \exp\!\left( -\frac{(x - \mu)^2}{2\sigma^2} \right),
\quad -\infty < x < \infty,
$$

with mean $\mu$ at the center and standard deviation $\sigma$ controlling the spread. There is no simple
closed form for its cdf, so we lean on two practical tools instead of integrating by hand.

The first is the **68–95–99.7 rule**: for any normal, about $68\%$ of the probability lies within $1$
standard deviation of the mean, about $95\%$ within $2$, and about $99.7\%$ within $3$. This rule of thumb
lets you sanity-check a normal probability in your head before trusting a table or a computer.

The second tool is **standardization**. Every normal is a shifted, rescaled copy of the single **standard
normal** $Z \sim \text{Normal}(0, 1)$. If $X \sim \text{Normal}(\mu, \sigma)$, then

$$
Z = \frac{X - \mu}{\sigma} \sim \text{Normal}(0, 1),
\qquad
P(X \le x) = \Phi\!\left( \frac{x - \mu}{\sigma} \right),
$$

where $\Phi$ is the standard-normal cdf. So you never need a separate table for each $\mu$ and $\sigma$:
convert your value to a $z$-score — a count of standard deviations from the mean — and read $\Phi(z)$ once.
A score $2$ standard deviations above the mean, for instance, has $\Phi(2) \approx 0.977$ below it, which
is just the upper end of the $95\%$ rule.

### Recognizing the model: a short checklist

When a continuous scenario lands in front of you, ask in order:

1. **Is every value in a fixed range equally likely, with no part favored?** → Uniform.
2. **Are you measuring a waiting time (or lifetime) where events arrive at a steady rate?** → Exponential
   (and recall its rate is the Poisson rate from Week 9).
3. **Is the quantity a measurement that clusters symmetrically around a typical value, with extremes rare?**
   → Normal.

As with the discrete catalogue, the checklist also tells you when *no* standard model fits cleanly: if the
rate is not constant, the exponential is only an approximation; if a quantity is strictly positive and
skewed, the normal (which allows any real value) may be a poor fit. Naming the limit of a model is part of
using it honestly.

## Worked examples

### Worked example — Maya's morning: the shuttle wait and the commute (recurring slice)

This is one connected morning in our recurring synthetic world; two continuous models appear back to back.
*(Synthetic scenario; numbers fixed for the course.)*

**The shuttle wait, as an exponential.** Shuttles arrive at Maya's stop at a steady rate of $\lambda = 4$
per hour — the same Poisson rate from Week 9 — so the time $T$ until the next one is exponential:

$$
T \sim \text{Exponential}(\lambda = 4 \text{ per hour}),
\qquad
f(t) = 4 e^{-4 t}, \quad t \ge 0,
$$

with time measured in hours. Its mean wait is

$$
E[T] = \frac{1}{\lambda} = \frac{1}{4}\ \text{hour} = 15\ \text{minutes},
$$

matching the intuition "one every fifteen minutes." Now: what is the chance she waits at most $15$ minutes
— at most $1/4$ hour? Use the cdf with $t = 1/4$, so that $\lambda t = 4 \times 1/4 = 1$:

$$
P(T \le 15\text{ min}) = F\!\left(\tfrac{1}{4}\right) = 1 - e^{-\lambda t} = 1 - e^{-1} \approx 0.632.
$$

So a little under two-thirds of the time her wait is at or below the *average* — a reminder that for a
right-skewed model the mean is not the median, and "at most the mean wait" is more likely than a coin flip.

**The commute, as a normal.** Once aboard, Maya's commute time $C$ (in minutes) clusters symmetrically
around a typical value with rare extremes, so we model it as normal:

$$
C \sim \text{Normal}(\mu = 22,\ \sigma = 5).
$$

What is the chance the commute takes at most $30$ minutes? Standardize $x = 30$:

$$
z = \frac{x - \mu}{\sigma} = \frac{30 - 22}{5} = \frac{8}{5} = 1.6,
\qquad
P(C \le 30) = \Phi(1.6) \approx 0.945.
$$

So about $94.5\%$ of mornings the commute is $30$ minutes or less. The $68$–$95$–$99.7$ rule confirms the
ballpark: $30$ is between $1\sigma$ ($27$) and $2\sigma$ ($32$) above the mean, so the cumulative
probability should sit between $0.84$ and $0.975$ — and $0.945$ does.

### Worked example — time until a component fails, as an exponential (transfer)

Now a new context with the same model. A small electronic component fails at a constant average rate of
$\lambda = 0.5$ failures per year, with no wear-in or wear-out trend — the assumption that makes the
**exponential** the right lifetime model. Let $L$ be its lifetime in years:

$$
L \sim \text{Exponential}(\lambda = 0.5 \text{ per year}),
\qquad
f(\ell) = 0.5\, e^{-0.5 \ell}, \quad \ell \ge 0,
$$

so the **mean** time to failure is $E[L] = 1/\lambda = 1/0.5 = 2$ years. What is the probability the
component survives past $4$ years? Use the complement of the cdf:

$$
P(L > 4) = 1 - F(4) = e^{-\lambda \ell} = e^{-0.5 \times 4} = e^{-2} \approx 0.135.
$$

About a $13.5\%$ chance it lasts beyond $4$ years. The **memoryless** property gives a striking corollary:
a component that has *already* survived $4$ years has the same probability of surviving another $2$ years
as a brand-new one — $P(L > 6 \mid L > 4) = P(L > 2) = e^{-1} \approx 0.368$ — because the exponential
keeps no memory of time already elapsed. This is the same machinery as Maya's shuttle wait, applied to
hardware instead of buses. *(Synthetic scenario; numbers fixed for the course.)*

### Seeing it in software (shown, not run)

Software returns all of these probabilities directly. Below, R's `p*` functions give cdf values
$P(X \le x)$ and `d*` functions give density heights $f(x)$. The code is **shown to read**, not executed in
this diagnostic.

```{r}
#| label: continuous-models-cdf
#| eval: false

# Shuttle wait T ~ Exponential(rate = 4 per hour); 15 minutes = 0.25 hour
pexp(0.25, rate = 4)            # P(T <= 15 min) = 1 - e^-1 ~ 0.632

# Commute C ~ Normal(mean = 22, sd = 5) minutes
pnorm(30, mean = 22, sd = 5)    # P(C <= 30) ~ 0.945
pnorm(1.6)                      # Phi(1.6): standardized, same answer ~ 0.945

# Component lifetime L ~ Exponential(rate = 0.5 per year)
1 - pexp(4, rate = 0.5)         # P(L > 4) = e^-2 ~ 0.135

# Density heights (NOT probabilities): f(x) for plotting the curves
dnorm(22, mean = 22, sd = 5)    # peak of the commute density, at the mean
dexp(0,  rate = 4)              # exponential density height at t = 0, equals the rate
dunif(0.1, min = 0, max = 0.25) # flat height 1/(b - a) = 4 over the interval
```

```{r}
#| label: continuous-models-simulate
#| eval: false

# Optional: simulate to confirm the two recurring-slice probabilities by frequency.
set.seed(35003)
waits <- rexp(100000, rate = 4)              # simulated shuttle waits, in hours
mean(waits <= 0.25)                          # empirical P(T <= 15 min), ~ 0.632

commutes <- rnorm(100000, mean = 22, sd = 5) # simulated commute times, in minutes
mean(commutes <= 30)                         # empirical P(C <= 30), ~ 0.945
```

The simulation chunk uses `set.seed(35003)` so it reproduces exactly when later executed; the empirical
fractions should land near the exact values above. *(Synthetic; seed set.)*

## A common mistake

The most frequent error this week is **mixing up the exponential's rate and mean**. We parameterize the
exponential by its **rate** $\lambda$, so $\text{Exponential}(4)$ has mean $1/4$, not $4$. In R this is the
`rate =` argument; passing the mean where a rate belongs (or vice versa) silently gives the reciprocal of
the answer you want. Always ask: is the number I have an *events-per-unit-time* (a rate) or a *time-per-
event* (a mean)? For the shuttle, $4$ per hour is the rate; $15$ minutes is the mean.

A second slip is **treating a density value $f(x)$ as a probability**. The bell curve's height
$dnorm(22, 22, 5) \approx 0.080$ is *not* the chance the commute equals exactly $22$ minutes — that
probability is zero, exactly as Week 10 insisted. Density is probability *per unit length*; only the
**area** between two values is a probability. And for the uniform, the flat height can even exceed $1$ (it
is $4$ on $[0, 0.25]$) precisely because it is a density, not a probability.

## Low-stakes self-checks (ungraded)

These are for your own practice — ungraded, no submission, just a way to test whether the recognition skill
is sticking. Try them before peeking at the reasoning.

1. A bus is equally likely to arrive at any moment in the next $10$ minutes. Which model fits, and what is
   the probability it arrives in the first $3$ minutes? (Uniform$(0, 10)$; $P = 3/10 = 0.3$, a proportion of
   the interval.)
2. For Maya's commute $C \sim \text{Normal}(22, 5)$, roughly what fraction of mornings is the commute
   between $17$ and $27$ minutes, *without* a calculator? (Those are $\mu \pm 1\sigma$, so about $68\%$ by
   the rule.)
3. A help desk receives calls at a steady rate of $\lambda = 6$ per hour. What model describes the wait
   until the next call, and what is the mean wait? (Exponential$(6)$; mean $1/6$ hour $= 10$ minutes.)
4. Why is $P(C = 22)$ equal to $0$ for the normal commute, even though $22$ is its mean and the most likely
   *region*? (It is continuous: any single exact value has zero area, so zero probability; only intervals
   carry probability.)
5. A component's lifetime is $\text{Exponential}(0.5)$ per year. Without recomputing, is the probability it
   survives its *next* year larger if it is brand new or if it has already lasted $3$ years? (The same —
   memoryless; past survival does not change the remaining-wait distribution.)

You can check the probability statements by simulation using the shown `rexp`/`rnorm`/`runif` chunks above:
draw many values and compare the empirical fraction to the exact answer.

## Reading and source pointer

This week tracks **Grinstead & Snell, Chapter 5 — Important Distributions**, specifically the *continuous*
distributions (uniform, exponential, and normal), for the catalogue of named models and the
parameterizations we use. The standardization $Z = (X - \mu)/\sigma$ and the $68$–$95$–$99.7$ framing for
reading normal probabilities are supported by the review of continuous distributions and the normal in
**MIT OCW 18.05**. Our parameterizations follow the course ledger: $\text{Uniform}(a, b)$;
$\text{Exponential}(\lambda)$ as a **rate** with mean $1/\lambda$; $\text{Normal}(\mu, \sigma)$ as **mean
and standard deviation**. The means quoted here are stated results carried from the expectation work in
Week 8 and the density/area machinery in Week 10, not re-derived. *These notes are the course's own
synthesis, grounded in but not copied from the sources.* All scenario data are synthetic, with seeds set in
any simulation.

## Formula-verification status

`verified: false`. The densities, cdfs, means, and numeric probabilities on this page are **drafted and
provisional**: the course **math gate is BLOCKED** pending human sign-off. Treat every formula and number
here — the exponential cdf $1 - e^{-\lambda t}$, the mean $1/\lambda$, the standardization, and the worked
values $1 - e^{-1} \approx 0.632$ and $\Phi(1.6) \approx 0.945$ — as **provisional pending human
verification**, not as confirmed results.

## Public vs. graded

These notes, the examples, and the practice here are **public and ungraded** — study material only. No
graded prompts, answer keys, rubrics, point values, or due dates appear on this site. Graded checkpoints,
quizzes, homework, labs, the midterm, the project, and the final live in **Blackboard (the LMS)**, which is
authoritative for due dates, submissions, and grades. If this page and Blackboard ever disagree, follow
Blackboard.

## Looking ahead

So far every random variable — discrete or continuous — has been a *single* quantity. Next week we let two
vary together: **joint distributions and dependence**. We return to the rain-and-lateness pair from Weeks
3–4, now as random variables $X$ (rain) and $Y$ (late), and measure how they move together with
**covariance** and **correlation**. After that, Week 13 brings the normal back in force: the Central Limit
Theorem explains *why* the bell curve we used for Maya's commute is so common — it is what the average of
many independent pieces looks like, no matter their individual shape.

## See also

- [Notation glossary](../resources/notation-glossary.qmd) — the symbols, from $f(x)$ and $F(x)$ to
  $\lambda$, $\mu$, $\sigma$, and $\Phi$, used here.
- [Distribution reference](../resources/distribution-reference.qmd) — a one-page summary of the uniform,
  exponential, and normal: parameters, densities, and means.
- [Lab 13 — Law of large numbers and CLT](../labs/lab-13-law-of-large-numbers-and-clt.qmd) — the
  simulation lab where the normal returns as the shape of an average.
- [Syllabus](../syllabus.qmd) — course structure, calendar, and where graded work actually lives.
